---
title: "REACT Analysis"
date: August 15, 2026
created by: Victoria Chu
---

# Setup
```{r setup-clear-data}

rm(list = ls())
graphics.off()

```

## Packages and paths

```{r setup-libraries, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Project paths
DATA_DIR <- file.path("/Users/victoriachu/Documents/Research/REACT/FinalData") #../DataFiles

# Statistical modeling and publication tables
library(gtsummary)
library(geepack)
library(glmmTMB)
library(car)
library(ggpubr)

# Core data wrangling and plotting
library(tidyverse)
library(lubridate)
library(scales)
library(forcats)
library(ggrepel)
library(patchwork)
library(gridExtra)
library(viridis)

# Network and phylogenetic analysis
library(igraph)
library(ggtree)
library(ggraph)
library(tidygraph)
library(ggtreeExtra)
library(ggnewscale)
library(vegan)
library(ape)
library(ComplexHeatmap)
library(circlize)
```

```{r setup-theme}
theme_mine <- function(base_size = 12, base_family = "") {
  theme_bw(base_size = base_size, base_family = base_family) %+replace%
    theme(
      strip.background  = element_blank(),
      strip.text.x      = element_text(size = 18),
      strip.text.y      = element_text(size = 18),
      axis.text.x       = element_text(size = 14),
      axis.text.y       = element_text(size = 14, hjust = 1),
      axis.ticks        = element_line(colour = "black"),
      axis.title.x      = element_text(size = 16),
      axis.title.y      = element_text(size = 16, angle = 90),
      panel.background  = element_blank(),
      panel.border      = element_blank(),
      panel.grid.major  = element_blank(),
      panel.grid.minor  = element_blank(),
      panel.spacing     = unit(1.0, "lines"),
      plot.background   = element_blank(),
      plot.margin       = unit(c(0.5, 0.5, 0.5, 0.5), "lines"),
      axis.line         = element_line(colour = "black")
    )
}

theme_set(theme_mine())
```

```{r setup-risk-parameters}

# Shared time scale used by the risk-factor analysis.
STUDY_START <- as.Date("2023-05-07")

week_dates <- tibble(
  week     = paste0("W", 1:12),
  week_num = 1:12,
  w_start  = STUDY_START + weeks(0:11),
  w_end    = STUDY_START + weeks(0:11) + days(6)
)

week_daily_lookup <- week_dates %>%
  rowwise() %>%
  mutate(date = list(seq(w_start, w_end, by = "day"))) %>%
  unnest(date) %>%
  select(date, week, week_num)

```

## Data input

```{r setup-input-data}

patients_all <- read.csv(file.path(DATA_DIR, "patients.csv"), na = c("NA", ""))
patients <- patients_all %>% filter(!(nPID %in% c("P08", "P14")))

samples <- read.csv(file.path(DATA_DIR, "samples.csv"), na = c("NA", ""))

rooms <- read.csv(file.path(DATA_DIR, "rooms.csv"), na = c("NA", ""))
providers <- read.csv(file.path(DATA_DIR, "providers.csv"), na = c("NA", ""))

micro_cx <- read.csv(file.path(DATA_DIR, "micro_cultures_filt.csv"), na = c("NA", ""))

mNGS_samples <- read.csv(file.path(DATA_DIR, "mNGS_samples.csv"), na = c("NA", ""))

mNGS_instrain <- read.csv(file.path(DATA_DIR, "SOI_instrain_newthreshold.csv"), na = c("NA", ""))
instrain_old <- read.csv(file.path(DATA_DIR, "SOI_instrain_oldthreshold.csv"), na = c("NA", ""))

MAGs_SOI <- read.csv(file.path(DATA_DIR, "MAGs_SOI.csv"), na = c("NA", ""))
MAGs_all <- read.csv(file.path(DATA_DIR, "MAGs_all.csv"), na = c("NA", ""))
MAGs <- read.csv(file.path(DATA_DIR, "MAGs_all_filt.csv"), na = c("NA", ""))
MAGs_amr <- read.csv(file.path(DATA_DIR, "MAGs_amr.csv"), na = c("NA", ""))

SSS <- read.csv(file.path(DATA_DIR, "strain_sharing_samples.csv"), na = c("NA", ""))

# Risk-factor preprocessing derived from the shared room and provider tables.

rooms_raw <- rooms %>%
  mutate(
    AdmDate = as.Date(AdmDate),
    DcDate  = as.Date(DcDate),
    Room    = ifelse(Room == 520, paste(Room, "-", Bed, sep = ""), Room)
  )

providers_raw <- providers %>%
  mutate(
    event_date = as.Date(event_date),
    room = case_when(
      room == "SCU 2" ~ "520-2",
      room == "SCU 3" ~ "520-3",
      room == "SCU 4" ~ "520-4",
      room == "SCU 5" ~ "520-5",
      TRUE            ~ room
    )
  )

```


# Analysis

## 1. Study enrollment

```{r main_Demographics}

non_donor_pt <- patients %>% filter(Study_Group!="Donor")

summary(non_donor_pt$Age)

table(non_donor_pt$Gender)
table(non_donor_pt$Race)
table(non_donor_pt$Ethnicity)
table(non_donor_pt$Diabetes)
table(non_donor_pt$Hemodialysis)


# Summarize all collected samples
all_samples_summ <- samples %>% filter(Collected==1 & Study_Group!="Donor")

#number participants per sampling week
tbl_summary(
  all_samples_summ,
  by = nPID,
  include = c(nTID, nPID),
  statistic = all_categorical() ~ "{n} ({p}%)"
) %>%
  add_overall()

tbl_summary(
  all_samples_summ,
  by = Source,
  include = c(nTID, Source),
  statistic = all_categorical() ~ "{n} ({p}%)"
) %>%
  add_overall()

#number of weeks for SNT patients
samples_data <- samples %>% filter(Collected==1 & Study_Group=="SNT") %>%
  select(nPID, nTID) %>% distinct() %>%
  group_by(nPID) %>%
  mutate(weeks_num = n()) %>%
  select(nPID, weeks_num) %>%
  distinct()

summary(samples_data$weeks_num)


#number samples per source
tbl_summary(samples %>% filter(Collected==1) %>% select(Source),
  statistic = all_categorical() ~ "{n} ({p}%)"
)

table(all_samples_summ$Source)


```

## 2. WGS characterization

### 2.1 WGS summary

```{r main_WGS summary}

WGS <- micro_cx %>% filter(WGS_yn==1)

strain <- as.data.frame(unique(WGS$WGS_bact_strain))%>%
  mutate(
    species = str_extract(`unique(WGS$WGS_bact_strain)`, "^.+(?=_)"),   # everything before last underscore
    number  = str_extract(`unique(WGS$WGS_bact_strain)`, "[^_]+$")       # everything after last underscore
  ) %>% distinct() %>%
  select(species, number)

```

### 2.2 Supplementary Table 1

```{r Supplementary Table 1}

table(WGS$MDRO) # number of isolates
table(strain$species) # number of strains

```


### 2.3 Supplementary Figure 1

```{r WGS phylo tree}
# Build radial phylogenetic tree from multiple Newick files
# Load species trees
tree_files <- list.files(
  path = "../REACTTree",
  pattern = "\\.(nwk|newick)$",
  full.names = TRUE,
  recursive = TRUE
)

cat("Found", length(tree_files), "Newick files:\n")
print(basename(tree_files))

# Define helper to extract species names from filenames
get_species_label <- function(file) {
  tools::file_path_sans_ext(basename(file))
}

# Define the desired species order
desired_order <- c(
  "Kpneumo_tree",
  "Klebspp_tree",
  "Enterobacter_spp_phylo_tree_0326",
  "Citrobacter_tree",
  "ecoli_phylo_tree_0326",
  "Pmirabilis_tree",
  "Paeruginosa_tree",
  "Smaltophilia_tree",
  "Acinetobacter_tree",
  "Efaecalis_tree",
  "Cperfingens_tree",
  "Cdiff_tree",
  "Efaecium_tree"
)

# Read and label species trees
species_trees <- lapply(tree_files, function(file) {
  tr <- read.tree(file)
  if (inherits(tr, "multiPhylo")) tr <- tr[[1]]
  species <- get_species_label(file)
  tr$tip.label <- paste(species, tr$tip.label, sep = "_")
  attr(tr, "species_name") <- species
  tr
})

# Reorder species trees to the desired order
species_names <- sapply(species_trees, function(tr) attr(tr, "species_name"))

# Check that all requested species are present
missing <- setdiff(desired_order, species_names)
if (length(missing) > 0) warning("Missing species: ", paste(missing, collapse = ", "))

# Reorder species trees
species_trees <- species_trees[match(desired_order, species_names)]
species_names <- sapply(species_trees, function(tr) attr(tr, "species_name"))  # update

# Build scaffold tree with one tip per species
scaffold <- stree(length(species_names), tip.label = species_names, type = "left")

# Graft each species tree onto the scaffold
combined_tree <- scaffold
for (tr in species_trees) {
  species     <- attr(tr, "species_name")
  graft_point <- which(combined_tree$tip.label == species)
  combined_tree <- bind.tree(combined_tree, tr, where = graft_point, position = 0)
}

# Remove the original scaffold tips, which are now internal nodes
combined_tree <- drop.tip(combined_tree, species_names)

cat("\nCombined tree summary:\n")
cat("  Tips:         ", length(combined_tree$tip.label), "\n")
cat("  Internal nodes:", combined_tree$Nnode, "\n")

# Prepare tip metadata
tip_data <- data.frame(
  label   = combined_tree$tip.label,
  species = sub("_.*", "", combined_tree$tip.label),
  stringsAsFactors = FALSE
)

tip_data <- tip_data %>%
  mutate(temp_label = tolower(label)) %>%
  mutate(temp_label = str_remove(temp_label, "tree_|ecoli_phylo_tree_0326_|enterobacter_spp_phylo_tree_0326_")) %>%
  mutate(temp_label = str_remove(temp_label, "acinetobacter|cdiff|cperfingens|citrobacter|ecoli|efaecium|efaecalis|paeruginosa|kpneumo|klebspp|pmirabilis|enterobacter|smaltophilia")) %>%
  mutate(temp_label = str_remove(temp_label, "_cdiff_extracted|_efaecalis_extracted|_efaecium_extracted|_kpneumoniae_extracted|_pmirabilis_extracted")) %>%
  mutate(temp_label = ifelse(species!="ecoli" & species!="Enterobacter", gsub("^[^_]*_|_combined$", "", temp_label), str_remove(temp_label,"_combined"))) %>%
  mutate(temp_label = case_when(temp_label=="p23220-s008_apps0018-2igvre05-22-23-1of1_s8" & species =="Efaecium" ~ "p23220-s008_apps0018-2igvre05-22-23-1of2_s8",
                                temp_label=="p23220-s008_apps0018-2igvre05-22-23-1of1_s8" & species =="Pmirabilis" ~ "p23220-s008_apps0018-2igvre05-22-23-2of2_s8",
                                TRUE ~ temp_label)) %>%
  left_join(micro_cx %>% select(WGS_sample_name, WGS_bact_strain, Source), by=c("temp_label" = "WGS_sample_name")) %>%
  mutate(Source = ifelse(Source=="env", "environmental", "patient"),
         strain_num = as.integer(str_extract(WGS_bact_strain, "(?<=_)\\d+$"))) %>%
  mutate(Source = ifelse(is.na(Source), "reference genome", Source)) %>%
  mutate(species = case_when(WGS_bact_strain=="C. perfringens_1" ~ "C. perfringens",
                             WGS_bact_strain=="C. perfringens_2" ~ "C. perfringens",
                             TRUE ~ species))

# Create display names for the legend
species_map <- c(
  "Acinetobacter"        = "Acinetobacter pitii",
  "Cdiff"        = "Clostridioides difficile",
  "Cperfingens"  = "Clostridium perfringens",
  "C. perfringens"  = "Clostridium perfringens",
  "Citrobacter"  = "Citrobacter freundii",
  "ecoli"        = "Escherichia coli",
  "Efaecium"     = "Enterococcus faecium",
  "Efaecalis"    = "Enterococcus faecalis",
  "Paeruginosa"  = "Pseudomonas aeruginosa",
  "Kpneumo"      = "Klebsiella pneumoniae",
  "Klebspp"      = "Klebsiella spp.",
  "Pmirabilis"    = "Proteus mirabilis",
  "Enterobacter" = "Enterobacter cloacae complex",
  "Smaltophilia" = "Stenotrophomonas maltophilia"
)

tip_data$species_pretty <- species_map[tip_data$species]
tip_data$species_pretty[is.na(tip_data$species_pretty)] <-
  tip_data$species[is.na(tip_data$species_pretty)]

#changing the klebsiella spp. group
tip_data <- tip_data %>%
  mutate(species_pretty = case_when(
    WGS_bact_strain == "K. aerogenes_1" ~ "Klebsiella aerogenes",
    WGS_bact_strain =="K. michiganensis_1" ~ "Klebsiella michiganensis",
    label == "Klebspp_tree_reference" ~ "Klebsiella michiganensis",
    TRUE ~ species_pretty  # keep original for everything else
  )) %>%
  mutate(species_pretty = case_when(grepl("E. hormaechei",WGS_bact_strain) ~ "Enterobacter hormaechei",
                                    grepl("E. kobei",WGS_bact_strain) ~ "Enterobacter kobei",
                                    label=="Enterobacter_spp_phylo_tree_0326_reference" ~ "Enterobacter hormaechei",
    TRUE ~ species_pretty  # keep original for everything else
  ))


### Isolates sequenced but absent from Newick files
tip_data %>%
  mutate(temp_label = tolower(label)) %>%
  mutate(check=1) %>%
  mutate(temp_label = gsub("^[^_]*_|_combined$", "", temp_label)) %>%
  full_join(micro_cx %>% select(WGS_sample_name, MDRO), by=c("temp_label" = "WGS_sample_name")) %>%
  filter(is.na(check) & !is.na(temp_label))

# Define color palettes
unique_pretty <- unique(tip_data$species_pretty)
n_species     <- length(unique_pretty)

WGS_colors<-c("#8B4789","#D7CE9FFF", "#B9AA97FF", "#2C5F2D","#4F6980FF","#849DB1FF",  "#A2CEAAFF",
 "#B66353FF","#FBB04EFF","#F47942FF", "#C9A84C", "#8E7B9B","#4A7C9E", "#5B8C85", "#638B66FF")


# Define strain colors


# Define source colors
source_levels <- sort(unique(na.omit(tip_data$Source)))
source_palette <- c("#E8967A", "#7EB8B2","gray75")
names(source_palette) <- source_levels

tip_data$Source <- factor(tip_data$Source, levels = source_levels)

p <- ggtree(
    combined_tree,
    layout = "circular",
    color  = "gray40",
    size   = 0.2
  ) %<+% tip_data

tip_order <- rev(p$data$label[!is.na(p$data$label)])

# Set factor levels to match tree order
# Order species by first appearance in the tree
species_order <- tip_data$species_pretty[match(tip_order, tip_data$label)]
species_levels <- c("Klebsiella pneumoniae", "Klebsiella michiganensis", "Klebsiella aerogenes", "Enterobacter hormaechei", "Enterobacter kobei", "Citrobacter freundii", "Escherichia coli", "Proteus mirabilis", "Pseudomonas aeruginosa", "Stenotrophomonas maltophilia", "Acinetobacter pitii",  "Enterococcus faecalis",  "Clostridium perfringens", "Clostridioides difficile","Enterococcus faecium")

tip_data$species_pretty <- factor(tip_data$species_pretty, levels = species_levels)

# Order sources using the same tree-based approach
source_order <- tip_data$Source[match(tip_order, tip_data$label)]
source_levels <- unique(na.omit(source_order))
tip_data$Source <- factor(tip_data$Source, levels = source_levels)

# Alternate source colors when strain assignments change
current_strain <- NULL
toggle <- TRUE
alt_color_map <- c()

for (i in seq_along(tip_order)) {
  lbl <- tip_order[i]
  idx <- which(tip_data$label == lbl)
  if (length(idx) == 0) next

  s <- tip_data$strain_num[idx]
  s_char <- ifelse(is.na(s), NA_character_, as.character(s))
  if (length(s_char) == 0 || is.na(s_char)) {
    alt_color_map[lbl] <- NA
    next
  }

  if (is.null(current_strain) || !identical(s_char, current_strain)) {
    toggle <- !toggle                # flip on each new strain
    current_strain <- s_char
  }
  alt_color_map[lbl] <- ifelse(toggle, "A", "B")
}

tip_data$strain_alt <- alt_color_map[tip_data$label]

# Count samples represented by each category
species_counts <- tip_data %>%
  count(species_pretty) %>%
  deframe()  # converts to named vector

source_counts <- tip_data %>%
  count(Source) %>%
  deframe()

# Rebuild the plot with counts in the labels
p <- ggtree(
    combined_tree,
    layout = "circular",
    color  = "gray40",
    size   = 0.2
  ) %<+% tip_data +
  geom_tippoint(
    aes(color = species_pretty),
    size = 2
  ) +
  scale_color_manual(
    name   = "Species (including reference)",
    values = WGS_colors,
    labels = function(x) paste0(x, " (n=", species_counts[x], ")"),
    na.translate = FALSE
  ) +

  # Inner ring: source
  geom_fruit(
    geom = geom_tile,
    mapping = aes(y = label, fill = Source),
    width  = 0.95,
    offset = 0.08
  ) +
  scale_fill_manual(
    name   = "Source",
    values = source_palette,
    labels = function(x) paste0(x, " (n=", source_counts[x], ")"),
    na.value = "transparent",
    na.translate = FALSE
  ) +

  new_scale_fill() +

  # Outer ring: alternating strain colors
  geom_fruit(
    geom = geom_tile,
    mapping = aes(y = label, fill = strain_alt),
    width  = 0.8,
    offset = 0.05
  ) +
  scale_fill_manual(
    name   = "Strain groups\n(alternating colors)",
    values = c("A" = "#3D3D5C", "B" = "#C4A882"),
    labels = c("A" = "",
               "B" = ""),
    na.value = "transparent",
    na.translate = FALSE,
    guide  = guide_legend(nrow = 1)
  ) +
  xlim(NA, max(p$data$x, na.rm = TRUE) * 1.3) +  # expand right side by 30%

  theme_void() +
  theme(
    legend.position = c(1, 0.5),
    legend.text = element_text(size = 6),
    legend.title = element_text(size = 8),
    legend.key.size = unit(0.4, "cm"),
    legend.box.spacing = unit(0.1, "cm")
  )

print(p)


```

### 2.4 WGS strain-sharing events

```{r WGS strain sharing events}

# Assign patients to WGS strains
strain_assignments <- micro_cx %>%
  filter(!is.na(WGS_t_cluster_strain)) %>%
  distinct(WGS_t_cluster_strain, nPID)

# Estimate possible transmission events assuming one facility introduction per strain cluster
strain_events <- strain_assignments %>%
  count(WGS_t_cluster_strain, name = "n_patients") %>%
  mutate(strain_sharing_events = n_patients - 1)

# Summarize overall strain-sharing events
total_events <- sum(strain_events$strain_sharing_events)
n_clusters <- nrow(strain_events)
unique_patients <- n_distinct(strain_assignments$nPID)


```


## 3. mNGS characterization

```{r mNGS sample metrics}

nrow(mNGS_samples)

mNGS_samples %>% summarise(n_ind = n_distinct(nPID))
num_samples_ind <- as.data.frame(table(mNGS_samples$nPID))
summary(num_samples_ind$Freq)


```


## 4. MAG characterization

### 4.1 MAG metrics

```{r MAG summary}

# MAG domain summary
tbl_summary(MAGs_all,
  include = c(domain)
)

# Percentage of MAGs with >=90% completeness
(nrow(MAGs_all %>% filter(SCG_completeness >= 90)) / nrow(MAGs_all)) * 100

# Percentage of MAGs with <500 contigs
(nrow(MAGs_all %>% filter(contigs <500)) / nrow(MAGs_all)) * 100

```


### 4.2 Supplementary Figure 2

```{r main_MAG richness}

# Figure 2A: MAG completeness distribution
completeness <- ggplot(MAGs_all, aes(x = SCG_completeness)) +
  geom_density(fill = "paleturquoise4", alpha = 0.5, color=NA) +
  labs(
    x = "SCG Completeness (%)",
    y = "Density"
  ) +
  scale_x_continuous(expand = c(0,0), limits = c(0, NA)) +
  scale_y_continuous(expand = c(0,0), limits = c(0, NA)) +
  theme(plot.margin = unit(c(0.5, 2, 0.5, 0.5), "lines")) +
  theme_classic() +
  theme(
    plot.margin = unit(c(0.5, 2, 0.5, 0.5), "lines"),
    axis.title = element_text(size = 10),
    axis.text = element_text(size = 9)
  )


# Figure 2B: Contig count distribution
contigs <- ggplot(MAGs_all, aes(x = contigs)) +
  geom_density(fill = "paleturquoise4", alpha = 0.5, color=NA) +
  labs(
    x = "Contigs per MAG",
    y = "Density"
  ) +
  scale_x_continuous(expand = c(0,0), limits = c(0, NA)) +
  scale_y_continuous(expand = c(0,0), limits = c(0, NA)) +
  theme_classic() +
  theme(
    axis.title = element_text(size = 10),
    axis.text = element_text(size = 9)
  )

# Figure 2C: MAG richness versus sequencing depth

# Scale total reads to millions
mNGS_samples_nbm <- mNGS_samples %>%
  mutate(total_10mil_reads = total_reads / 1e6)


# Fit the negative binomial model
nb_model <- glm.nb(
  MAG_count ~ total_10mil_reads + SDI_microbes,
  data = mNGS_samples_nbm
)

# Extract rate ratios and 95% confidence intervals
exp_coef <- exp(coef(nb_model))
ci <- exp(confint(nb_model))
results <- cbind(Rate_Ratio = exp_coef, CI_lower = ci[,1], CI_upper = ci[,2])
results


# Median SDI for visualization
median_SDI <- median(mNGS_samples_nbm$SDI_microbes)

# Prediction grid for plotting
newdat <- data.frame(
  total_10mil_reads = seq(min(mNGS_samples_nbm$total_10mil_reads),
                        max(mNGS_samples_nbm$total_10mil_reads),
                        length.out = 100),
  SDI_microbes = median_SDI
)

# Predicted MAGs and 95% CI
pred <- predict(nb_model, newdata = newdat, type = "link", se.fit = TRUE)
newdat <- newdat %>%
  mutate(
    fit = exp(pred$fit),
    lower = exp(pred$fit - 1.96 * pred$se.fit),
    upper = exp(pred$fit + 1.96 * pred$se.fit)
  )

summary(nb_model)

# Extract IRR and 95% CI
IRR <- exp(coef(nb_model))
CI  <- exp(confint(nb_model))  # profile likelihood CI

irr_table <- data.frame(
  Variable = names(IRR),
  IRR = IRR,
  CI_lower = CI[,1],
  CI_upper = CI[,2]
)

summary(nb_model)

irr_table


eq_label <- paste0(
  "log(E[MAGs]) = β0 + β1(reads/10M) + β2(SDI)"
)


reads_plot <- ggplot(mNGS_samples_nbm, aes(x = total_10mil_reads, y = MAG_count)) +
  geom_point(aes(color = Study_Group == "Donor"), alpha = 0.6, size = 2.5) +

  geom_line(data = newdat, aes(x = total_10mil_reads, y = fit),
            color = "black", size = 1, inherit.aes = FALSE) +

  geom_ribbon(data = newdat,
              aes(x = total_10mil_reads, ymin = lower, ymax = upper),
              alpha = 0.2, fill = "grey70", inherit.aes = FALSE) +

  scale_color_manual(
    values = c("FALSE" = "grey60", "TRUE" = "#FC8D62"),
    labels = c("Perirectal", "Stool"),
    name = "Sample type"
  ) +

  annotate(
    "text",
    x = Inf, y = Inf,
    label = "log(E[MAGs]) == beta[0] + beta[1]*(reads/10*M) + beta[2]*SDI",
    parse = TRUE,
    hjust = 1.3, vjust = 5.1, size = 3.5
  ) +

  labs(
    x = "Total reads (per 10 million)",
    y = "MAG richness"#,
  ) +

  scale_x_continuous(expand = expansion(mult = c(0, 0))) +  # 0% left, 5% right
  scale_y_continuous(expand = expansion(mult = c(0, 0))) +  # 0% bottom, 5% top
  coord_cartesian(clip = "off") +  # Allow plotting outside panel

  theme_classic() +
  theme(legend.position = c(0.5, 0.99),           # centered horizontally, top
    legend.justification = c(0.5, 1),         # anchor at top-center of legend
    legend.direction = "horizontal",          # horizontal layout
    plot.title = element_text(hjust = 0.5)
  )  +
  theme_classic() +
  theme(
    axis.title = element_text(size = 11),
    axis.text = element_text(size = 10),
    legend.text = element_text(size = 9),
    legend.title = element_text(size = 9),
    legend.position = c(0.5, 0.99),
    legend.justification = c(0.5, 1),
    legend.direction = "horizontal"
  )

# Prepare the model results table.
# Extract model p-values.
p_values <- coef(summary(nb_model))[, "Pr(>|z|)"]  # p-values from the z-test

# Add p-values to the IRR table.
irr_table <- irr_table %>%
  mutate(`P-value` = p_values[Variable])

# Format the IRR and 95% CI for display.
irr_table_edited <- irr_table %>%
  mutate(
    IRR = round(IRR, 1),
    CI_lower = round(CI_lower, 1),
    CI_upper = round(CI_upper, 1),
    IRR = sprintf("%.1f", IRR),
    CI_lower = sprintf("%.1f", CI_lower),
    CI_upper = sprintf("%.1f", CI_upper),
    `95% CI` = paste0("(", CI_lower, " - ", CI_upper, ")"),
    `P-value` = sprintf("%.1e", `P-value`)) %>%
  filter(Variable!="(Intercept)") %>%
  mutate(Variable = case_when(Variable=="total_10mil_reads" ~ "reads/10M",
                              Variable=="SDI_microbes" ~ "SDI"),
         ) %>%
  select(Variable, IRR, `95% CI`, `P-value`)

colnames(irr_table_edited) <- c("Variable", "IRR", "95% CI", "p-value")

# Convert the results table to a compact table grob.
table_grob <- tableGrob(
  irr_table_edited,
  rows = NULL,  # or set to rownames(irr_table_edited) if you want row names
  theme = ttheme_minimal(
    base_size = 8,  # smaller font size
    core = list(fg_params = list(cex = 1.3)),  # core cells font size
    colhead = list(fg_params = list(cex = 1.3))  # column header font size
  )
)

right_col <- reads_plot / plot_spacer() /  wrap_elements(table_grob, ignore_tag = TRUE) +
  plot_layout(heights = c(6,0.1, 1))

layout <- (completeness / contigs) | right_col
layout <- layout +
  plot_layout(widths = c(1, 2))

layout_annotated <- layout +
  plot_annotation(tag_levels = 'a') &
  theme(plot.tag = element_text(face = 'bold', size = 12))

layout_annotated

```

### 4.3 Supplementary Table 2

```{r Supplementary Table 2}

# All samples
tbl_summary(mNGS_samples,
  include = c(total_reads, SDI_microbes, MAG_count)
)

tbl_summary(MAGs_all,
  include = c(SCG_completeness, contigs, size)
)

#### 4.3.1 Perirectal samples
mNGS_pr <- mNGS_samples %>% filter(Source=="PR")
MAGs_pr <- MAGs_all %>% filter(Source=="PR")

tbl_summary(mNGS_pr,
  include = c(total_reads, SDI_microbes, MAG_count)
)

tbl_summary(MAGs_pr,
  include = c(SCG_completeness, contigs, size)
)

#### 4.3.2 Stool samples
mNGS_stool <- mNGS_samples %>% filter(Source=="stool")
MAGs_stool <- MAGs_all %>% filter(Source=="stool")

tbl_summary(mNGS_stool,
  include = c(total_reads, SDI_microbes, MAG_count)
)

tbl_summary(MAGs_stool,
  include = c(SCG_completeness, contigs, size)
)

```

### 4.4 Supplementary Figure 3

```{r AMR plot}
# Define colors used in Supplementary Figure 3
col_bar  <- "paleturquoise4"   # blue  – unique SGBs (bars)
col_line <- "coral"   # amber – total AMR genes (line)

#### 4.4.1 Figure A: SGBs and AMR genes by antibiotic class
keep_classes <- c(
  "Aminoglycoside", "Chloramphenicol", "Beta-lactam", "Fluoroquinolone",
  "Glycopeptide", "MLS", "Oxazolidinone", "Rifampin", "Sulfonamide",
  "Tetracycline", "Colistin", "Lipopeptide", "Fosfomycin", "Nitrofurantoin",
  "Multi-drug efflux pump"
)

fig_a_data <- MAGs_amr |>
  filter(abx_class %in% keep_classes) |>
  group_by(abx_class) |>
  summarise(
    unique_SGBs = n_distinct(MAG_derep_num, na.rm = TRUE),
    total_genes  = n(),
    .groups = "drop"
  ) |>
  arrange(desc(total_genes), desc(unique_SGBs)) |>
  mutate(abx_class = fct_inorder(abx_class))

scale_factor <- max(fig_a_data$total_genes) / max(fig_a_data$unique_SGBs)

# Plot Figure A.
fig_a <- ggplot(fig_a_data, aes(x = abx_class)) +
  geom_col(
    aes(y = total_genes),
    fill  = col_bar,
    alpha = 0.85,
    width = 0.7
  ) +
  scale_y_continuous(
    name     = "Number of MAG-based ARGs",
    breaks   = scales::breaks_pretty(n = 6),
    limits = c(0, 1400),
    expand = c(0,0)
  ) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 14)) +
  labs(
    x       = "\nAntibiotic class"
  ) +
  theme_bw(base_size = 11) +
  theme(
    axis.text.x        = element_text(angle = 90, hjust = 1, size = 8),
    axis.title.y  = element_text(size = 10),
    panel.grid.minor   = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption       = element_text(size = 8, colour = "grey50", hjust = 0),
    plot.margin        = margin(8, 12, 8, 8)
  )

print(fig_a)

#### 4.4.2 Figure B: Priority AMR genes by organism

# Define priority genes; category controls point color and gene_label controls the y-axis.
priority_genes <- tribble(
  ~gene_pattern,    ~gene_label,    ~category,
  # Carbapenemases
  "^KPC-3$",        "KPC-3",        "Carbapenemase",
  # ESBLs
  "^CTX-M-15$",       "CTX-M-15",     "ESBL",
  "^SHV-7$",        "SHV-7",        "ESBL",
  "^SHV-28$",       "SHV-28",       "ESBL",
  "^SHV-32$",       "SHV-32",       "ESBL",
  "^SHV-50$",       "SHV-50",       "ESBL",
  # Vancomycin resistance
  "^vanA$",           "vanA",         "Vancomycin resistance",
  "^vanB$",           "vanB",         "Vancomycin resistance",
  # Colistin
  "mcr",            "mcr",          "Colistin resistance",
  # Fluoroquinolone
  "^QnrB1$",        "QnrB1",        "Fluoroquinolone resistance",
  "^QnrS1$",        "QnrS1",        "Fluoroquinolone resistance"
)

detect_hits <- function(df, pattern, label, category) {
  df |>
    filter(str_detect(gene_name, regex(pattern, ignore_case = TRUE))) |>
    mutate(gene_label = label,
           category   = category)
}

fig_b_raw <- pmap_dfr(
  priority_genes,
   ~ detect_hits(MAGs_amr, ..1, ..2, ..3)
)

fig_b_raw <- fig_b_raw |>
  distinct(gene_label, category, MAG_derep_num, species)

cat_order <- c(
  "Carbapenemase",
  "ESBL",
  "Vancomycin resistance",
  "Colistin resistance",
  "Fluoroquinolone resistance"
)
gene_order <- c(
  "KPC-3", "CTX-M-15",
  "SHV-7", "SHV-28", "SHV-32", "SHV-50",
  "vanA", "vanB",
  "mcr",
  "QnrB1", "QnrS1"
)

fig_b_raw <- fig_b_raw |>
  mutate(
    y_label    = gene_label,   # just the gene name — no drug class in brackets
    category   = factor(category, levels = cat_order),
    gene_label = factor(gene_label, levels = gene_order)
  ) |>
  arrange(category, gene_label) |>
  mutate(y_label = fct_inorder(y_label))

cat_colours <- c(
  "Carbapenemase" = "#A32D2D",
  "ESBL"                              = "#F47942FF",
  "Vancomycin resistance"             = "paleturquoise4",
  "Colistin resistance"               = "black", #not actually detected
  "Fluoroquinolone resistance"        = "#8B4789"
)

# Match y-axis text colors to gene categories
y_label_order    <- levels(fig_b_raw$y_label)
y_label_category <- fig_b_raw |>
  distinct(y_label, category) |>
  arrange(match(y_label, y_label_order))
axis_text_colours <- rev(cat_colours[as.character(y_label_category$category)])

# Denominator: total distinct MAG IDs per species across all samples.
mags_per_species <- MAGs_amr |>
  distinct(species, MAG_ID) |>
  count(species, name = "total_mags")

# Numerator: distinct MAG IDs carrying each priority gene per species.
detect_hits_mag <- function(df, pattern, label, category) {
  df |>
    filter(str_detect(gene_name, regex(pattern, ignore_case = TRUE))) |>
    distinct(species, MAG_ID) |>
    mutate(gene_label = label,
           category   = category)
}

fig_b_raw_mag <- pmap_dfr(
  priority_genes,
  ~ detect_hits_mag(MAGs_amr, ..1, ..2, ..3)
)

# Create a key for joining y-axis labels.
y_label_key <- fig_b_raw |>
  distinct(gene_label, category, y_label)

fig_b_plot <- fig_b_raw_mag |>
  group_by(gene_label, category, species) |>
  summarise(mags_with_gene = n_distinct(MAG_ID), .groups = "drop") |>
  left_join(mags_per_species, by = "species") |>
  left_join(y_label_key, by = c("gene_label", "category")) |>
  mutate(
    prop    = mags_with_gene / total_mags,
    # clean GTDB clade suffixes (e.g. Enterococcus_B → Enterococcus)
    species_clean = str_replace(species, "_[A-Z](?= |$)", ""),
    # x-axis label: "Species name (n=XX)" where XX is the denominator
    x_label = paste0(species_clean, " (n=", total_mags, ")")
  )

# Order x-axis labels alphabetically by cleaned species name
x_order <- fig_b_plot |>
  distinct(x_label, species_clean) |>
  arrange(species_clean) |>
  pull(x_label)
fig_b_plot <- fig_b_plot |>
  mutate(
    x_label      = factor(x_label, levels = x_order),
    prop_label   = mags_with_gene
  )

# Plot Figure B.
fig_b <- ggplot(fig_b_plot,
                aes(x = x_label, y = y_label, colour = category, size = prop)) +
  geom_point(shape = 19, alpha = 0.85) +
  geom_text(
    aes(label = prop_label),
    colour = "white", size = 2.5, fontface = "bold"
  ) +
  scale_colour_manual(
    values = cat_colours,
    name   = "Resistance category"
  ) +
  scale_size_continuous(
    name   = "Proportion of MAGs with ARG detected",
    range  = c(2, 8),
    limits = c(0, 1),
    breaks = c(0.25, 0.5, 0.75, 1),
    labels = c("25%", "50%", "75%", "100%")
  ) +
  scale_x_discrete(position = "bottom") +
  scale_y_discrete(limits = rev(y_label_order)) +
  labs(
    x       = "\nBacterial species (n = MAG detections)",
    y       = "ARG\n"
  ) +
  theme_bw(base_size = 11) +
  theme(
    axis.text.x      = element_text(angle = 40, hjust = 1, size = 9,
                                    face = "italic"),
    axis.text.y      = element_text(size = 9,
                                    colour = axis_text_colours),
    axis.title       = element_text(size = 10),
    panel.grid.major = element_line(colour = "grey88", linewidth = 0.4),
    panel.grid.minor = element_blank(),
    legend.position  = "right",
    legend.title     = element_text(size = 9),
    legend.text      = element_text(size = 9),
    plot.caption     = element_text(size = 8, colour = "grey50", hjust = 0),
    plot.margin      = margin(8, 12, 8, 8),
    legend.justification = "left",
    legend.box.just = "left"
  ) +
  guides(
    colour = guide_legend(override.aes = list(size = 4), order = 1, nrow = 4),
    size   = guide_legend(order = 2, nrow = 4)
  )

print(fig_b)


# Extract the legend, then remove it from the main plot.
fig_b_legend <- cowplot::get_legend(fig_b)
fig_b_no_legend <- fig_b + theme(legend.position = "none")

# Assemble the final figure: Figure A on top, Figure B and legend below.
# Keep the legend in a dedicated panel.
AMR_plot <-
  fig_a /
  ((fig_b_no_legend | wrap_elements(fig_b_legend)) + plot_layout(widths = c(1.5, 1)))+
  plot_layout(heights = c(1, 1.5)) +
  plot_annotation(tag_levels = list(c("a", "b"))) &
  theme(plot.tag = element_text(face = "bold", size = 14))

AMR_plot

```


## 5. SGB characterization

```{r}

SGBs <- MAGs_all %>% filter(!is.na(MAG_derep_num)) %>%
  left_join(patients %>% select(nPID, Donor), by="nPID") %>%
  mutate(Donor = ifelse(Donor=="MEPSD0008", "D01",
                        ifelse(Donor=="MEPSD0015", "D02",
                               ifelse(Donor=="MEPSD0015, MEPSD0008", "Both",NA_character_))))

# Unique MAGs
max(SGBs$MAG_derep_num)

```


### 5.1 Supplementary Figure 4

```{r SGB bacterial phylo tree}

taxonomy_clean <- SGBs %>%
  mutate(phylum = ifelse(phylum=="Desulfobacterota_I", "Desulfobacterota", phylum)) %>%
  mutate(phylum = ifelse(is.na(phylum), "Unclassified", phylum),
         class  = ifelse(is.na(class),  paste(phylum,"class"),  class),
         order  = ifelse(is.na(order),  paste(class,"order"),   order),
         family = ifelse(is.na(family), paste(order,"family"),  family),
         genus  = ifelse(is.na(genus),  paste(family,"genus"),  genus),
         species= ifelse(is.na(species),paste(genus,"species"), species)) %>%
  mutate(species = ifelse(phylum=="Unclassified" | domain=="Unclassified", MAG_derep, species)) %>%
  mutate(genus = ifelse(phylum=="Unclassified" | domain=="Unclassified", MAG_derep, genus)) %>%
  mutate(family = ifelse(phylum=="Unclassified" | domain=="Unclassified", MAG_derep, family)) %>%
  mutate(order = ifelse(phylum=="Unclassified" | domain=="Unclassified", MAG_derep, order)) %>%
  mutate(class = ifelse(phylum=="Unclassified" | domain=="Unclassified", MAG_derep, class)) %>%
  mutate(
    phylum_id  = paste0("p__", phylum),
    class_id   = paste0(phylum_id,  "|c__", class),
    order_id   = paste0(class_id,   "|o__", order),
    family_id  = paste0(order_id,   "|f__", family),
    genus_id   = paste0(family_id,  "|g__", genus),
    species_id = paste0(genus_id,   "|s__", species),
    domain_id  = paste0("d__Bacteria"))

length(unique(taxonomy_clean$genus))


species_per_phylum <- taxonomy_clean %>%
  select(species_id, phylum) %>%   # species = terminal nodes
  distinct() %>%                   # one row per species
  count(phylum, name = "n_species") %>%
  arrange(desc(n_species))

levels <- c("domain_id","phylum_id","class_id",
            "order_id","family_id","genus_id","species_id")

edges <- map_dfr(seq_along(levels)[-1], function(i){
  taxonomy_clean %>%
    dplyr::select(from = all_of(levels[i - 1]),
           to   = all_of(levels[i])) %>%
    distinct()
})

tax_tree <- graph_from_data_frame(edges, directed = TRUE)
root_node <- which(V(tax_tree)$name == "domain_id")

tax_tree <- tax_tree %>%
  as_tbl_graph() %>%
  activate(nodes) %>%
  mutate(sort_key = name) %>%   # alphabetical
  arrange(sort_key)

tg <- tax_tree %>%
  activate(nodes) %>%
  mutate(
    phylum = stringr::str_extract(name, "p__[^|]+"),
    phylum = gsub("^p__", "", phylum),
    is_domain = grepl("^d__", name)
  )

phylum_levels <- tg %>%
  activate(nodes) %>%
  as_tibble() %>%
  filter(!is.na(phylum)) %>%
  distinct(phylum) %>%
  pull(phylum)

phylum_levels <- c(
  sort(phylum_levels[phylum_levels != "Unclassified"]),
  "Unclassified"
)

tg <- tg %>%
  activate(nodes) %>%
  mutate(
    phylum = factor(phylum, levels = phylum_levels)
  )

tg <- tg %>%
  activate(nodes) %>%
  arrange(
    phylum,                    # order by phylum_levels
    !node_is_leaf(),           # draw internal nodes first
    name                       # stable tie-breaker
  )

my_colors<-c("#4F6980FF","#849DB1FF", "#A2CEAAFF", "#638B66FF", "#BFBB60FF", "#F47942FF",
"#FBB04EFF", "#B66353FF","#D7CE9FFF", "#B9AA97FF", "black")


# Ensure all factor levels are represented in the intended order.
phylum_counts_df <- tibble(phylum = phylum_levels) %>%
  left_join(species_per_phylum, by = "phylum") %>%
  replace_na(list(n_species = 0))

# Create legend labels in the same order as phylum_levels.
phylum_labels <- paste0(phylum_counts_df$phylum, " (n=", phylum_counts_df$n_species, ")")

total_n <- sum(species_per_phylum$n_species)


tree <- ggraph(tg, layout = "dendrogram", circular = TRUE) +
  geom_edge_diagonal(color = "gray80", width = 0.2) +
  geom_node_point(
    aes(color = phylum),
    size = 1.5,
    na.rm = TRUE
  ) +
  scale_color_manual(
    name = paste0("Bacterial phylum (N=", total_n, ")"),
    values = my_colors,
    labels = phylum_labels,
    na.translate = FALSE
  ) +
  theme_void() +
  theme(
    legend.position = "right",
    legend.text = element_text(size = 8)
  )

tree

```

### 5.2 Figure 2

```{r Figure 2 Data}

mag_to_sgb <- MAGs %>%
  group_by(mNGS_sample_name) %>%
  summarise(
    total_MAGs = n(),
    high_quality_MAGs = sum(!is.na(MAG_derep_num)),
    recovery_rate = high_quality_MAGs / total_MAGs * 100
  ) %>%
  left_join(mNGS_samples %>% select(mNGS_sample_name, nPID, Study_Group, Source),
            by = "mNGS_sample_name")


```

#### 5.2.1 Figure 2B

```{r Figure 2B}

p1 <- ggplot(mag_to_sgb, aes(x = total_MAGs, y = high_quality_MAGs)) +
  geom_point(alpha = 0.5, size = 1, color="grey60") +
  geom_smooth(method = "lm", se = FALSE, color="black", size=0.5) +
  theme_classic() +
  labs(x = "MAG richness", y = "SGB richness") +
  theme(legend.position = "bottom", legend.title = element_blank()) +
  coord_cartesian(clip = "off") +
  scale_x_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.02))) +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.02))) +
        stat_regline_equation(label.x=29, label.y=7, size=3)

```


#### 5.2.2 Figure 2C

```{r Figure 2C}
p2 <- ggplot(mag_to_sgb, aes(x = recovery_rate)) +
  geom_histogram(bins = 20, fill = "grey60", alpha = 0.7) +
  theme_classic() +
  labs(x = "MAG to SGB recovery rate (%)", y = "Number of samples") +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.02)))

```


#### 5.2.3 Figure 2D

```{r Figure 2D}
p3 <- ggplot(mag_to_sgb %>%
               mutate (Source = ifelse(Source=="PR","Perirectal", "Stool")),
             aes(x = Source, y = recovery_rate, color = Source)) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.5) +
  geom_boxplot(outlier.shape = NA, width = 0.35, fill=NA, color="black", size = 0.5) +
  scale_color_manual(values = c("Stool" = "coral",
                                 "Perirectal" = "grey60")) +
  theme_classic() +
  labs(x = "", y = "MAG to SGB\nrecovery rate (%)") +
  theme(legend.position = "none") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.02)), breaks = c(0, 25, 50, 75, 100)) +
  coord_cartesian(
    ylim = c(0, NA),
    clip = "off"
  )

# Wilcoxon rank-sum test for recovery rate
wilcox.test(recovery_rate ~ Source,
            data = mag_to_sgb %>%
              mutate(Source = ifelse(Source == "PR", "Perirectal", "Stool")))

```

#### 5.2.4 Figure 2E

```{r Figure 2E}

detection_freq <- SGBs %>%
  group_by(MAG_derep_num) %>%
  summarise(
    n_samples = n_distinct(mNGS_sample_name),
    n_participants = n_distinct(nPID),
    pct_samples = n_samples / n_distinct(mNGS_samples$mNGS_sample_name) * 100
  )


sgb_sharing <- detection_freq %>%
  mutate(sharing = case_when(
    n_participants == 1 ~ "1 person",
    n_participants <= 5 ~ "2-5 people",
    n_participants <= 15 ~ "6-15 people",
    TRUE ~ ">15 people"
  ))

sgb_sharing_summary <- sgb_sharing %>%
  count(sharing) %>%
  mutate(sharing = factor(sharing, levels = c(">15 people",
                                               "6-15 people",
                                               "2-5 people",
                                               "1 person")))

p4 <- ggplot(sgb_sharing_summary, aes(x = sharing, y = n, fill = sharing)) +
  geom_col(width = 0.85) +
  geom_text(aes(label = n), hjust = -0.2, size = 3) +
  scale_fill_manual(values = c("1 person" = "grey60",
                                "2-5 people" = "#A2CEAAFF",
                                "6-15 people" = "paleturquoise4",
                                ">15 people" = "#2F6B6D")) +
  coord_flip() +
  theme_classic() +
  theme(legend.position = "none") +
  labs(x = "", y = "Number of SGBs") +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.2)))

```


#### 5.2.5 Figure 2F

```{r Figure 2F}
samples_with_MAGs <- SGBs %>%
  distinct(mNGS_sample_name)

zero_samples <- setdiff(mNGS_samples$mNGS_sample_name, samples_with_MAGs$mNGS_sample_name)

species_mat <- SGBs %>%
  distinct(mNGS_sample_name, MAG_derep) %>%
  mutate(present = 1) %>%
  pivot_wider(
    names_from = MAG_derep,
    values_from = present,
    values_fill = 0
  ) %>%
  column_to_rownames("mNGS_sample_name") %>%
  as.matrix()

jaccard_dist <- vegdist(species_mat, method = "jaccard", binary = TRUE)

pcoa_res <- pcoa(jaccard_dist)

pcoa_df <- data.frame(
  sample = rownames(species_mat),
  PC1 = pcoa_res$vectors[,1],
  PC2 = pcoa_res$vectors[,2]
)

n_excluded <- length(zero_samples)

pcoa_df2 <- pcoa_df %>%
  left_join(mNGS_samples, by = c("sample" = "mNGS_sample_name")) %>%
  left_join(samples %>% select(nSID, Room) %>% distinct(),
            by="nSID") %>%
  mutate(Population = ifelse(nPID=="D01", "Donor 1",
                             ifelse(nPID=="D02","Donor 2" ,"LTAC residents"))) %>%
  mutate(TID_v2 = ifelse(Study_Group=="Donor","Donor",
                         nTID)) %>%
  mutate(TID_v2 = factor(TID_v2, levels = c("Donor", paste0("W", 1:12))))  %>%
  group_by(nPID) %>%
  arrange(nPID, TID_v2) %>%  # Sort by individual AND time
  mutate(relative_timepoint = row_number(),
         max_timepoints = n(),
         alpha_scaled = relative_timepoint / max_timepoints, # scale 0 to 1 within individual
         color_group = ifelse(n() == 1, "Single sample", as.character(nPID))) %>%
  ungroup() %>%
  arrange(nPID, TID_v2)

Jac_dis1 <- ggplot(pcoa_df2, aes(PC1, PC2, color=Population)) +
  geom_point(size = 2, alpha = 0.8) +
  theme_classic() +
  labs(
    x = paste0("Axis 1 (", round(pcoa_res$values$Relative_eig[1] * 100, 1), "%)"),
    y = paste0("Axis 2 (", round(pcoa_res$values$Relative_eig[2] * 100, 1), "%)")
  ) +
  scale_x_continuous(limits = c(-0.8, 0.25)) +
  scale_color_manual(values = c("Donor 1" = "paleturquoise4",
                                "Donor 2" = "#A2CEAAFF",
                                "LTAC residents" = "coral")) + 
  theme(axis.text = element_blank(), 
        axis.ticks = element_blank(),
        legend.position = c(0.02, 0.98),
        legend.background = element_rect(fill = NA, color = NA),
        legend.justification = c(0, 1))

```

#### 5.2.6 Figure 2G

```{r Figure 2G}
dist_df <- as.matrix(jaccard_dist) %>%
  as.data.frame() %>%
  rownames_to_column("sample1") %>%
  pivot_longer(-sample1, names_to = "sample2", values_to = "distance") %>%
  left_join(pcoa_df2 %>% select(sample, nPID), by = c("sample1" = "sample")) %>%
  rename(nPID1 = nPID) %>%
  left_join(pcoa_df2 %>% select(sample, nPID), by = c("sample2" = "sample")) %>%
  rename(nPID2 = nPID) %>%
  filter(sample1 < sample2) %>%  # avoid duplicates
  filter(!sample1 %in% zero_samples & !sample2 %in% zero_samples) %>%  # exclude zero-SGB samples
  mutate(comparison = ifelse(nPID1 == nPID2, "Within individual", "Between individuals"))

# Add SGB counts to the pairwise distance table.
dist_df <- dist_df %>%
  left_join(
    SGBs %>%
      group_by(mNGS_sample_name) %>%
      summarise(n_SGBs = n_distinct(MAG_derep)),
    by = c("sample1" = "mNGS_sample_name")
  ) %>%
  rename(n_SGBs1 = n_SGBs) %>%
  left_join(
    SGBs %>%
      group_by(mNGS_sample_name) %>%
      summarise(n_SGBs = n_distinct(MAG_derep)),
    by = c("sample2" = "mNGS_sample_name")
  ) %>%
  rename(n_SGBs2 = n_SGBs) %>%
  mutate(min_SGBs = pmin(n_SGBs1, n_SGBs2, na.rm = TRUE))

Jac_distance_horizontal <- ggplot(dist_df %>%
                                    mutate(comparison = ifelse(comparison == "Within individual", "Within\nindividual", "Between\nindividuals")),
                                  aes(x = distance, y = comparison, color = comparison)) +
  geom_jitter(data = dist_df %>% filter(distance != 0) %>%
                mutate(comparison = ifelse(comparison == "Within individual", "Within\nindividual", "Between\nindividuals")),
              height = 0.1, size = 1, alpha = 0.3) +
  geom_point(data = dist_df %>% filter(distance == 0) %>%
               mutate(comparison = ifelse(comparison == "Within individual", "Within\nindividual", "Between\nindividuals")),
             size = 1, alpha = 0.3) +
  geom_boxplot(outlier.shape = NA, width = 0.35, color = "black", fill = NA, size = 0.5) +
  scale_color_manual(values = c("Within\nindividual" = "coral",
                                 "Between\nindividuals" = "paleturquoise4")) +
  scale_y_discrete(limits = rev) +  
  theme_classic() +
  theme(legend.position = "none") +
  coord_cartesian(clip = "off") +
  labs(
    x = "Jaccard dissimilarity",
    y = ""
  ) +
  scale_x_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.02)))

```

#### 5.2.7 Figure 2H

```{r Figure 2H}
Jac_scatter <- ggplot(dist_df, aes(x = min_SGBs, y = distance, color = comparison, fill=comparison)) +
  geom_point(alpha = 0.3, size = 1) +
  geom_smooth(method = "loess", se = TRUE, alpha = 0.2, size=0.5) +
  scale_color_manual(values = c("Within individual" = "coral",
                                 "Between individuals" = "paleturquoise4")) +
  scale_fill_manual(values = c("Within individual" = "coral",
                                "Between individuals" = "paleturquoise4")) +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.02))) +
  scale_x_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.02))) +
  coord_cartesian(clip = "off") +
  theme_classic() +
  labs(
    x = "SGB richness of the SGB-poorer sample",
    y = "Jaccard dissimilarity",
    color = ""
  ) +
  guides(fill="none", color = guide_legend(override.aes = list(fill = NA))) + 
  theme(legend.position = c(0.6, 0.2),
        legend.background = element_blank())

```

#### 5.2.8 Figure 2 layout

```{r Figure 2 Layout}

bottom_row1 <- (p4 | Jac_dis1) + plot_layout(widths = c(1, 1.5))
bottom_row2 <- (Jac_distance_horizontal | Jac_scatter) + plot_layout(widths = c(1, 1.5))

bottom_layout <- bottom_row1 / bottom_row2

top_layout <- (p1 | p2 | p3) + plot_layout(widths = c(1, 1, 1))

layout_annotated <- top_layout / bottom_layout +
  plot_layout(heights = c(1, 4)) +
  plot_annotation(tag_levels = list(letters[2:26])) &
  theme(plot.tag = element_text(face = 'bold', size = 12))

layout_annotated

```


### 5.3 Supplementary Figure 5

```{r top SGBs heatmap}


# Step 1: Select the most prevalent SGBs.
top_sgbs <- SGBs %>%
  group_by(MAG_derep) %>%
  summarise(n_persons = n_distinct(nPID), .groups = "drop") %>%
  arrange(desc(n_persons)) %>%
  filter(n_persons >= 6) %>%
  pull(MAG_derep)

# Step 2: Count samples per participant.
total_samples_per_pid <- mNGS_samples %>%
  group_by(nPID) %>%
  summarise(total_samples = n_distinct(mNGS_sample_name))

all_pids <- total_samples_per_pid %>% pull(nPID)

# Step 3: Create the SGB percentage matrix.
heatmap_data <- SGBs %>%
  filter(MAG_derep %in% top_sgbs) %>%
  group_by(nPID, MAG_derep) %>%
  summarise(n_samples_detected = n_distinct(mNGS_sample_name), .groups = "drop") %>%
  left_join(total_samples_per_pid, by = "nPID") %>%
  mutate(pct_detected = (n_samples_detected / total_samples) * 100) %>%
  select(nPID, MAG_derep, pct_detected) %>%
  complete(nPID = all_pids, MAG_derep = top_sgbs, fill = list(pct_detected = 0)) %>%
  pivot_wider(names_from = nPID, values_from = pct_detected, values_fill = 0) %>%
  column_to_rownames("MAG_derep") %>%
  as.matrix()

mag_label_map <- data.frame(MAG_derep = rownames(heatmap_data)) %>%
  left_join(
    SGBs %>% distinct(MAG_derep, MAG_short = str_remove(MAG_derep, "_[^_]*$")),
    by = "MAG_derep"
  ) %>%
  mutate(MAG_short = ifelse(is.na(MAG_short) | MAG_short == "NA",
                            "Unclassified", MAG_short)) %>%
  mutate(MAG_short = gsub("_", " ", MAG_short))

row_labels <- mag_label_map$MAG_short

# Verify annotation lengths match the heatmap dimensions.
print(paste("Number of rows in heatmap_data:", nrow(heatmap_data)))
print(paste("Length of row_labels:", length(row_labels)))

# Step 5: Create the annotation data frame.
annotation_col <- total_samples_per_pid %>%
  left_join(
    SGBs %>% distinct(nPID, Donor),
    by = "nPID"
  ) %>%
  mutate(
    `Number of samples` = total_samples,
    Donor = case_when(Donor == "Both" ~ "D02, then D01",
                   is.na(Donor) ~ "None",
                   TRUE ~ Donor)
  ) %>%
  arrange(match(nPID, colnames(heatmap_data))) %>%
  column_to_rownames("nPID") %>%
  select(`Number of samples`, Donor)

# Step 6: Define ComplexHeatmap colors.
col_fun_samples <- colorRamp2(
  c(1, 7),
  c("white", "#7A3F33")
)

col_donor <- c(
  "D01" = "#D86621",
  "D02" = "#1A9C78",
  "D02, then D01" = "#5C4478",
  "None" = "white"  # or use "#FFFFFF"
)

ha_bottom <- HeatmapAnnotation(
  `Number of samples` = annotation_col$`Number of samples`,
  Donor = annotation_col$Donor,
  col = list(
    `Number of samples` = col_fun_samples,
    Donor = col_donor
  ),
  which = "column",
  annotation_name_side = "left",
  annotation_name_gp = gpar(fontsize = 6),
  simple_anno_size = unit(2, "mm"),
annotation_legend_param = list(
  `Number of samples` = list(
    at = c(1, 4, 7),
    labels = c("1", "4", "7"),
    title_gp = gpar(fontsize = 10, fontface = "bold"),
    labels_gp = gpar(fontsize = 8)
  ),
  Donor = list(
    title_gp = gpar(fontsize = 10, fontface = "bold"),
    labels_gp = gpar(fontsize = 8)
  )
)
)

# Step 8: Create the ComplexHeatmap.
ht <- Heatmap(
  heatmap_data,
  name = "% Detection",
  col = colorRampPalette(c("white", "#3A4F61"))(100),

  # Clustering
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  clustering_distance_rows = "euclidean",
  clustering_distance_columns = "euclidean",
  clustering_method_rows = "ward.D2",
  clustering_method_columns = "ward.D2",

  # Row settings
  row_labels = row_labels,
  show_row_names = TRUE,
  row_names_gp = gpar(fontsize = 6),
  row_dend_width = unit(10, "mm"),

  # Column settings
  show_column_names = TRUE,
  column_names_gp = gpar(fontsize = 6),
  column_dend_height = unit(7, "mm"),

  # Splitting heatmap

  # Annotations BELOW heatmap
  bottom_annotation = ha_bottom,

  # Add grid lines
  rect_gp = gpar(col = "gray90", lwd = 0.5),

  # Legend
  heatmap_legend_param = list(
    title = "% Detection",
    title_gp = gpar(fontsize = 10, fontface = "bold"),
    labels_gp = gpar(fontsize = 8)
  )
)

ht

```


### 5.4 Supplementary Figure 6

```{r Temporal Jaccard PCoA}

multi_pids <- unique(pcoa_df2$nPID[pcoa_df2$color_group != "Single sample"])

my_colors <- c("#4F6980FF","#849DB1FF", "#A2CEAAFF", "#638B66FF", "#BFBB60FF", "#F47942FF",
               "#FBB04EFF", "#B66353FF","#D7CE9FFF", "#B9AA97FF")

my_palette_21 <- colorRampPalette(my_colors)(21)

pid_colors <- setNames(my_palette_21, multi_pids)
color_values <- c("Single sample" = "grey60", pid_colors)


first_points <- pcoa_df2 %>%
  arrange(nPID, alpha_scaled) %>%   # replace with real time var if you have one
  group_by(nPID) %>%
  slice(1) %>%                      # keep first (earliest) sample
  ungroup() %>%
  filter(color_group != "Single sample")  # add this line

Jac_dis2 <- ggplot(pcoa_df2 %>% filter(color_group!="Single sample"), aes(PC1, PC2, color = color_group)) +
  geom_path(aes(group = nPID), alpha = 0.7, size=0.5,
            arrow = arrow(type = "closed", length = unit(0.3, "cm")),
            show.legend = FALSE) +  # Changed to geom_path
  geom_point(aes(alpha = alpha_scaled), size = 3) +
  geom_text(
    data = first_points,
    aes(PC1, PC2, label = nPID),
    size = 4,
    vjust = -0.8,
    fontface = "bold",
    show.legend = FALSE) +
  scale_color_manual(name=NULL,
    values = color_values) +
  scale_alpha_continuous(name = "Relative time point",
                         range = c(0.2, 1),
                         guide="none") +
  theme_classic() +
  theme(legend.position = "none") +
  labs(
          x = paste0("Axis 1 (", round(pcoa_res$values$Relative_eig[1] * 100, 1), "%)"),
    y = paste0("Axis 2 (", round(pcoa_res$values$Relative_eig[2] * 100, 1), "%)")
  ) +
  theme(legend.position = "right",
        axis.text = element_blank(),      # Remove axis labels
        axis.ticks = element_blank())   # Remove axis ticks

Jac_dis2


```


## 6. inStrain threshold and network analysis

### 6.1 inStrain setup

```{r network_instrain function}

plot_strain_network <- function(dataframe, mag_id_to_plot) {

  # Filter to specific MAG
  edges_mag <- dataframe %>%
    filter(MAG_ID == mag_id_to_plot) %>%
    select(ID_1, WGS_strain_num_1, ID_2, WGS_strain_num_2, Seq_1, Seq_2)

  # Step 1: Find all mNGS-WGS pairs that should be merged
  # (same ID, one is mNGS, one is WGS, and they're connected by an edge)
  mngs_wgs_pairs <- edges_mag %>%
    filter(ID_1 == ID_2,  # Same patient/timepoint/source
           Seq_1 != Seq_2,  # One is mNGS, one is WGS
           (Seq_1 == "WGS" | Seq_2 == "WGS"),  # At least one is WGS
           (Seq_1 == "mNGS" | Seq_2 == "mNGS"))  # At least one is mNGS

  # Create mapping: mNGS ID -> WGS ID (with strain number)
  mngs_to_wgs_map <- bind_rows(
    mngs_wgs_pairs %>%
      filter(Seq_1 == "mNGS", Seq_2 == "WGS") %>%
      transmute(mngs_id = ID_1, wgs_id = paste(ID_2, WGS_strain_num_2, sep = "_")),
    mngs_wgs_pairs %>%
      filter(Seq_1 == "WGS", Seq_2 == "mNGS") %>%
      transmute(mngs_id = ID_2, wgs_id = paste(ID_1, WGS_strain_num_1, sep = "_"))
  ) %>%
    distinct()

  # Step 2: Create unified IDs
  ## 1.7 WGS analysis always gets strain appended
  ## 1.2 mNGS and MAG characterization gets mapped to WGS ID if it shares strain, otherwise keeps original
  edges_mag <- edges_mag %>%
      left_join(mngs_to_wgs_map %>% rename(ID_1 = mngs_id, unified_ID_1_mapped = wgs_id),
            by = "ID_1") %>%
  left_join(mngs_to_wgs_map %>% rename(ID_2 = mngs_id, unified_ID_2_mapped = wgs_id),
            by = "ID_2") %>%
  mutate(
    unified_ID_1 = case_when(
      Seq_1 == "WGS" & !is.na(WGS_strain_num_1) ~ paste(ID_1, WGS_strain_num_1, sep = "_"),
      Seq_1 == "mNGS" & !is.na(unified_ID_1_mapped) ~ unified_ID_1_mapped,
      TRUE ~ ID_1
    ),
    unified_ID_2 = case_when(
      Seq_2 == "WGS" & !is.na(WGS_strain_num_2) ~ paste(ID_2, WGS_strain_num_2, sep = "_"),
      Seq_2 == "mNGS" & !is.na(unified_ID_2_mapped) ~ unified_ID_2_mapped,
      TRUE ~ ID_2
    )
  )

  # Remove self-loops (edges where unified_ID_1 == unified_ID_2)
  edges_mag <- edges_mag %>%
    filter(unified_ID_1 != unified_ID_2)

  # Check if we have any edges left
  if (nrow(edges_mag) == 0) {
    return(NULL)  # No network to plot
  }

  # Create graph with unified IDs
  g <- graph_from_data_frame(edges_mag[, c("unified_ID_1", "unified_ID_2")],
                             directed = FALSE)

  # Find connected components (strain clusters)
  clusters <- components(g)

  # Add cluster membership as vertex attribute
  V(g)$strain_cluster <- clusters$membership

  # Add patient ID
  V(g)$patient <- sub("_T.*$", "", V(g)$name)

  # Add sequencing type information
  seq_info <- bind_rows(
    edges_mag %>% select(unified_ID = unified_ID_1, seq_type = Seq_1),
    edges_mag %>% select(unified_ID = unified_ID_2, seq_type = Seq_2)
  ) %>%
    group_by(unified_ID) %>%
    summarise(
      has_mNGS = any(seq_type == "mNGS"),
      has_WGS = any(seq_type == "WGS"),
      seq_label = case_when(
        has_mNGS & has_WGS ~ "Both",
        has_mNGS ~ "mNGS",
        has_WGS ~ "WGS"
      ),
      .groups = "drop"
    )

  V(g)$seq_type <- seq_info$seq_label[match(V(g)$name, seq_info$unified_ID)]

  # Convert to tidygraph for ggraph
  tg <- as_tbl_graph(g)

  # Get species name for this MAG
  species_name <- unique(dataframe$species[dataframe$MAG_ID == mag_id_to_plot])

  # Plot
  p <- ggraph(tg, layout = 'fr') +
    geom_edge_link(aes(), alpha = 0.5, color = "gray50") +
    geom_node_point(aes(color = factor(strain_cluster),
                       shape = seq_type),
                   size = 3) +
    geom_node_text(aes(label = name),
                   repel = TRUE,
                   size = 2,
                   max.overlaps = Inf) +
    scale_color_manual(
      name = "Strain Cluster",
      values = rev(hcl.colors(14, "Earth"))  # Set to max you might need
      ) +
    scale_shape_manual(values = c("mNGS" = 16, "WGS" = 17, "Both" = 15),
                       name = "Sequencing Type") +
    labs(title = paste0("SGB: ", mag_id_to_plot),
         subtitle = paste0("Species: ", species_name)) +
    theme_graph() +
    theme(legend.position = "right",
          plot.subtitle = element_text(size = 11),
          plot.title.position = "plot")

  return(p)
}


```


### 6.2 Summary statistics

```{r summary stats}
# Summarize pairwise inStrain comparison statistics
summary_stats <- instrain_old %>%
  filter(shared_strain == TRUE & meets_min_scaffolds==TRUE) %>%
  summarise(
    #popANI
    popANI_Q1 = quantile(median_popANI, 0.25, na.rm = TRUE),
    popANI_median = median(median_popANI, na.rm = TRUE),
    popANI_Q3 = quantile(median_popANI, 0.75, na.rm = TRUE),

    #shared scaffold proportion
    prop_high_Q1 = quantile(prop_scaffolds_high, 0.25, na.rm = TRUE),
    prop_high_median = median(prop_scaffolds_high, na.rm = TRUE),
    prop_high_Q3 = quantile(prop_scaffolds_high, 0.75, na.rm = TRUE),

    #breadth
    prop_MAG_Q1 = quantile(prop_MAG_scaffolds, 0.25, na.rm = TRUE),
    prop_MAG_median = median(prop_MAG_scaffolds, na.rm = TRUE),
    prop_MAG_Q3 = quantile(prop_MAG_scaffolds, 0.75, na.rm = TRUE),
  )

print("Summary Statistics:")
print(summary_stats)

```


### 6.3 Original threshold analysis

```{r network_create_weighted_edges}

# Create weighted edges from strain comparison data
# Filter shared strains using minimum scaffold requirements
# Calculate similarity metrics and match to WGS strains

edges_weighted_old <- instrain_old %>%
  filter(shared_strain == TRUE & meets_min_scaffolds == TRUE) %>%
  mutate(
  nSID_1 = if_else(Seq_1 == "WGS" & !is.na(WGS_strain_num_1),
                   paste(ID_1, WGS_strain_num_1, sep = "_"),
                   paste(ID_1, sep = "_")),
  nSID_2 = if_else(Seq_2 == "WGS" & !is.na(WGS_strain_num_2),
                   paste(ID_2, WGS_strain_num_2, sep = "_"),
                   paste(ID_2, sep = "_"))) %>%
  select(MAG_ID, species, nSID_1, nSID_2, Seq_1, Seq_2,
         n_scaffolds_pair, median_popANI, prop_MAG_scaffolds, prop_scaffolds_high,
         WGS_bact_strain_1, WGS_bact_strain_2, WGS_strain_num_1, WGS_strain_num_2) %>%
  mutate(
    similarity = median_popANI * prop_scaffolds_high,
    distance = 1 - median_popANI,
    same_WGS_strain = !is.na(WGS_bact_strain_1) &
                      !is.na(WGS_bact_strain_2) &
                      WGS_bact_strain_1 == WGS_bact_strain_2
  )

```


### 6.4 Threshold grid search

```{r network_threshold_optimization}


# Define threshold combinations to test
threshold_grid <- expand.grid(
  min_popANI = c(0.99995, 0.99999, 0.999995, 0.999999),
  prop_high_thresh = c(0.8, 0.85, 0.9, 0.95, 1.0),
  prop_MAG_thresh = c(0.5, 0.6, 0.7)
)

# Helper function for threshold evaluation
mean_or_na <- function(x) {
  if (length(x) == 0) NA_real_ else mean(x)
}


# Function to test threshold combinations and calculate validation metrics
test_thresholds <- function(min_popANI, prop_high_t, prop_MAG_t) {
  max_dist <- 1 - min_popANI

  # Filter edges based on thresholds
  edges_filtered <- edges_weighted_old %>%
    filter(
      n_scaffolds_pair>=10,
      median_popANI >= min_popANI,
      prop_scaffolds_high >= prop_high_t,
      prop_MAG_scaffolds >= prop_MAG_t
    ) %>%
  select(MAG_ID, species, nSID_1, WGS_strain_num_1, nSID_2, WGS_strain_num_2,
         Seq_1, Seq_2, median_popANI, prop_MAG_scaffolds, prop_scaffolds_high) %>%
  mutate(
    similarity = median_popANI * prop_scaffolds_high,
    distance = 1 - median_popANI)


  n_edges_retained <- nrow(edges_filtered)

  # Return empty result if no edges pass threshold
  if (n_edges_retained == 0) {
    return(tibble(
      max_distance_from_1 = max_dist,
      min_popANI = min_popANI,
      prop_high_thresh = prop_high_t,
      prop_MAG_thresh = prop_MAG_t,
      n_edges = 0,
      pct_edges_retained = 0,
      n_total_clusters = 0,
      n_singletons = NA_integer_,
      longitudinal_consistency = NA_real_,
      WGS_strain_consistency = NA_real_,
      WGS_strain_purity = NA_real_,
      avg_modularity = NA_real_
    ))
  }

  # Build strain clusters for each MAG/species combination
clusters <- edges_filtered %>%
  group_by(MAG_ID, species) %>%
  group_modify(~ {
    edge_df <- .x %>% select(nSID_1, nSID_2)
    vertices <- unique(c(edge_df$nSID_1, edge_df$nSID_2))

    g <- graph_from_data_frame(edge_df, directed = FALSE, vertices = vertices)
    comp <- components(g)

    tibble(
      nSID = names(comp$membership),
      strain_cluster = paste0("C", comp$membership)
    )
  }) %>%
  ungroup() %>%
  mutate(WGS_strain_num = as.integer(str_extract(nSID, "[0-9]+$")),
           nSID = sub("^(([^_]*_){2}[^_]*)_.*", "\\1", nSID))

  # Calculate singleton count
  all_samples_filtered <- unique(c(edges_filtered$nSID_1, edges_filtered$nSID_2))
  samples_in_clusters <- unique(clusters$nSID)
  n_singletons <- length(setdiff(all_samples_filtered, samples_in_clusters))

  # Validate longitudinal consistency (same participant over time)
  validation_longitudinal <- clusters %>%
    mutate(nPID = sub("_W.*$", "", nSID)) %>%
    group_by(MAG_ID, species, nPID) %>%
    summarise(
      n_timepoints = n_distinct(nSID),
      n_clusters = n_distinct(strain_cluster),
      same_cluster = n_clusters == 1,
      .groups = "drop"
    ) %>%
    filter(n_timepoints > 1)

  # Build WGS strain mapping for each MAG/species
  wgs_mapping <- bind_rows(
    edges_filtered %>% select(MAG_ID, species, nSID = nSID_1, WGS_strain_num = WGS_strain_num_1),
    edges_filtered %>% select(MAG_ID, species, nSID = nSID_2, WGS_strain_num = WGS_strain_num_2)
  ) %>%
    distinct(MAG_ID, species, nSID, WGS_strain_num) %>%
    filter(!is.na(WGS_strain_num))

  # Join WGS labels to clusters
  clusters_with_WGS <- clusters %>%
    left_join(wgs_mapping, by = c("MAG_ID", "species", "nSID", "WGS_strain_num"))

  # Calculate WGS strain consistency (sensitivity: same WGS strain stays together)
  validation_WGS_sensitivity <- clusters_with_WGS %>%
    filter(!is.na(WGS_strain_num)) %>%
    group_by(MAG_ID, species, WGS_strain_num) %>%
    summarise(
      n_samples = n(),
      n_clusters = n_distinct(strain_cluster),
      same_cluster = n_clusters == 1,
      .groups = "drop"
    ) %>%
    filter(n_samples > 1)

  # Calculate WGS strain purity (specificity: each cluster contains only one WGS strain)
  validation_WGS_purity <- clusters_with_WGS %>%
    filter(!is.na(WGS_strain_num)) %>%
    group_by(MAG_ID, species, strain_cluster) %>%
    summarise(
      n_samples = n(),
      n_WGS_strains = n_distinct(WGS_strain_num),
      is_pure = n_WGS_strains == 1,
      .groups = "drop"
    ) %>%
    filter(n_samples > 1)

  # Calculate modularity for each MAG/species
  modularity_vals <- edges_filtered %>%
    group_by(MAG_ID, species) %>%
    group_modify(~ {
      edge_df <- .x %>% select(nSID_1, nSID_2, median_popANI, prop_scaffolds_high)

      if (nrow(edge_df) == 0) {
        return(tibble(modularity = NA_real_))
      }

      vertices <- unique(c(edge_df$nSID_1, edge_df$nSID_2))
      g <- graph_from_data_frame(
        edge_df[, c("nSID_1", "nSID_2")],
        directed = FALSE,
        vertices = vertices
      )

      # Weight edges by product of ANI and scaffold proportion
      E(g)$weight <- edge_df$median_popANI * edge_df$prop_scaffolds_high

      if (ecount(g) == 0 || vcount(g) <= 1) {
        return(tibble(modularity = NA_real_))
      }

      clust <- tryCatch(
        cluster_louvain(g, weights = E(g)$weight),
        error = function(e) NULL
      )

      if (is.null(clust)) {
        return(tibble(modularity = NA_real_))
      }

      tibble(modularity = modularity(clust))
    }) %>%
    ungroup()

  # Aggregate validation metrics
  long_consistency <- if (nrow(validation_longitudinal) > 0) {
    mean(validation_longitudinal$same_cluster)
  } else {
    NA_real_
  }

  WGS_sensitivity <- if (nrow(validation_WGS_sensitivity) > 0) {
    mean(validation_WGS_sensitivity$same_cluster)
  } else {
    NA_real_
  }

  WGS_purity <- if (nrow(validation_WGS_purity) > 0) {
    mean(validation_WGS_purity$is_pure)
  } else {
    NA_real_
  }

  # Count total clusters
  n_total_clusters <- clusters %>%
    distinct(MAG_ID, species, strain_cluster) %>%
    nrow()

  # Return results
  tibble(
    max_distance_from_1 = max_dist,
    min_popANI = min_popANI,
    prop_high_thresh = prop_high_t,
    prop_MAG_thresh = prop_MAG_t,
    n_edges = n_edges_retained,
    pct_edges_retained = (n_edges_retained / nrow(edges_weighted_old)) * 100,
    n_total_clusters = n_total_clusters,
    n_singletons = n_singletons,
    longitudinal_consistency = long_consistency,
    WGS_strain_consistency = WGS_sensitivity,
    WGS_strain_purity = WGS_purity,
    avg_modularity = mean(modularity_vals$modularity, na.rm = TRUE)
  )
}


```

#### 6.4.1 Supplementary Table 3

```{r Supplementary Table 3}

# Summarize threshold optimization results
threshold_results <- threshold_grid %>%
  pmap_dfr(~ test_thresholds(..1, ..2, ..3))
  
threshold_results

```

### 6.5 Figure 3

#### 6.5.1 Figure 3A

```{r network_threshold_visualization}

# Prepare data for visualization
candidates_plot <- threshold_results %>%
  filter(
    !is.na(WGS_strain_purity),
    !is.na(longitudinal_consistency),
    longitudinal_consistency >= 0.0,
    avg_modularity >= 0.3
  ) %>%
  mutate(
    prop_MAG_thresh = factor(prop_MAG_thresh),
    prop_high_lab = paste0("high=", prop_high_thresh)
  )  %>% 
  mutate(prop_high_thresh = factor(prop_high_thresh))

# Select recommended threshold
chosen <- candidates_plot %>%
  filter(prop_high_thresh == 0.95, prop_MAG_thresh == "0.5") %>%
  arrange(desc(pct_edges_retained)) %>%
  slice(1) %>%
  mutate(prop_high_thresh = factor(prop_high_thresh))

# Use fallback selection if filtering returns no rows
if (nrow(chosen) == 0) {
  chosen <- candidates_plot %>%
    arrange(desc(WGS_strain_purity), desc(pct_edges_retained)) %>%
    slice(1)
}

# Create threshold optimization plot
p_optimization <- ggplot(
  candidates_plot %>% filter(min_popANI==0.99995),
  aes(y = longitudinal_consistency, x = WGS_strain_purity)
) +
  # Main scatter plot
  geom_point(
    aes(color = prop_high_thresh, size = pct_edges_retained, shape = prop_MAG_thresh),
    alpha = 0.8,
    stroke = 0.3
  ) +
  scale_size_continuous(
    name = "% edges retained",
    range = c(1.5, 5),
    labels = function(x) paste0(round(x, 0), "%")
  ) +
  scale_shape_manual(
    name = "Breadth",
    values = c("0.5" = 16, "0.6" = 17, "0.7" = 15)
  ) +
  #scale threshold
  scale_color_manual(
  name = "Shared scaffold proportion",
    values = c("0.8" = "#B66353FF", "0.85" = "#F47942FF", "0.9" = "#A2CEAAFF", "0.95" = "#638B66FF", "1" = "#4F6980FF")
) +
  # Add labels for chosen thresholds
  geom_text_repel(
    data = bind_rows(
      chosen %>% mutate(opt = "A")
    ),
    aes(label = paste0(
      "inStrain metrics:\n   shared scaffold proportion=", prop_high_thresh,
      ",\n   breadth=", prop_MAG_thresh,
      ",\n   median popANI=", min_popANI,
      "\nResults:\n   WGS strain purity=", round(WGS_strain_purity, 2),
      "\n   WGS strain consistency=", round(WGS_strain_consistency, 2),
      "\n   longitudinal consistency=", round(longitudinal_consistency, 2)
    )),
    nudge_x = 0,
    nudge_y = 0.07,
    min.segment.length = 0.1,
    box.padding = 0.35,
    point.padding = 0.3,
    size = 2,
    hjust=0
  ) +
  labs(
    y = "longitudinal consistency\n(sensitivity)",
    x = "WGS strain purity (specificity)"
  ) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "none")

p_optimization

```

#### 6.5.2 Figure 3B

```{r new thresholds}

#new thresholds obtained from evaluating network plots and longitudinal similarities

instrain_new <- instrain_old %>%
  mutate(meets_prop_MAG = ifelse(prop_MAG_scaffolds >=0.5, TRUE, FALSE),
         meets_prop_high = ifelse(prop_scaffolds_high >=0.95, TRUE, FALSE)) %>%
  mutate(shared_strain = ifelse(meets_min_scaffolds == TRUE & meets_prop_MAG==TRUE & meets_prop_high==TRUE & median_popANI>=0.99995 & n_scaffolds_pair>=10, TRUE, FALSE)) %>%
  filter(shared_strain==TRUE)

instrain_new_df <- instrain_new %>%
  filter(species == "Pseudomonas aeruginosa") %>%
  select(MAG_ID, species, ID_1, WGS_strain_num_1, ID_2, WGS_strain_num_2, Seq_1, Seq_2)

mag_id <- "APPS0055PR_S_2_S113_concoct_bin45"

p <- plot_strain_network(
  instrain_new_df,
  mag_id
)

p

```

### 6.6 Louvain clustering

```{r network_prepare_data}

# Create weighted edges from strain comparison data
edges_weighted_new <- instrain_new %>%
  filter(shared_strain == TRUE) %>%
  select(MAG_ID, species, ID_1, WGS_strain_num_1, ID_2, WGS_strain_num_2,
         Seq_1, Seq_2, median_popANI, prop_MAG_scaffolds, prop_scaffolds_high) %>%
  mutate(
    similarity = median_popANI * prop_scaffolds_high,
    distance = 1 - median_popANI,
    nSID_1 = ifelse(!is.na(WGS_strain_num_1), paste(ID_1, WGS_strain_num_1, sep = "_"),ID_1),
    nSID_2 = ifelse(!is.na(WGS_strain_num_2), paste(ID_2, WGS_strain_num_2, sep = "_"),ID_2))

# Create sequencing-type lookup for unified IDs
seq_type_lookup <- bind_rows(
  edges_weighted_new %>% select(nSID = nSID_1, seq_type = Seq_1),
  edges_weighted_new %>% select(nSID = nSID_2, seq_type = Seq_2)
) %>%
  group_by(nSID) %>%
  summarise(
    has_mNGS = any(seq_type == "mNGS"),
    has_WGS = any(seq_type == "WGS"),
    seq_label = case_when(
      has_mNGS & has_WGS ~ "Both",
      has_mNGS ~ "mNGS",
      has_WGS ~ "WGS"
    ),
    .groups = "drop"
  )

strain_clusters_simple <- edges_weighted_new %>%
  group_by(MAG_ID, species) %>%
  group_modify(~ {
    edge_df <- .x %>% select(nSID_1, nSID_2)
    vertices <- unique(c(edge_df$nSID_1, edge_df$nSID_2))

    g <- graph_from_data_frame(edge_df, directed = FALSE, vertices = vertices)
    comp <- components(g)

    tibble(
      nSID = names(comp$membership),
      strain_cluster = paste0("C", comp$membership)
    )
  }) %>%
  ungroup()

strain_clusters_louvain <- edges_weighted_new %>%
  group_by(MAG_ID, species) %>%
  group_modify(~ {
    edge_df <- .x %>% select(nSID_1, nSID_2)
    vertices <- unique(c(edge_df$nSID_1, edge_df$nSID_2))

    g <- graph_from_data_frame(edge_df, directed = FALSE, vertices = vertices)
    E(g)$weight <- .x$median_popANI * .x$prop_scaffolds_high

    clust <- cluster_louvain(g, weights = E(g)$weight, resolution = 0.3)

    tibble(
      nSID = names(membership(clust)),
      strain_cluster = paste0("C", clust$membership),
      modularity = modularity(clust)
    )
  }) %>%
  ungroup()

```

#### 6.6.1 Comparison statistics

```{r network_Validation measures}

# Build strain clusters for each MAG/species combination

strain_clusters_simple_val <- strain_clusters_simple %>%
  mutate(WGS_strain_num = str_extract(nSID, "[0-9]+$"),
           nSID = sub("^(([^_]*_){2}[^_]*)_.*", "\\1", nSID))

strain_clusters_louvain_val <- strain_clusters_louvain %>%
  mutate(WGS_strain_num = str_extract(nSID, "[0-9]+$"),
           nSID = sub("^(([^_]*_){2}[^_]*)_.*", "\\1", nSID))

#function to evaluate the clusters +/- louvain
evaluate_clusters <- function(clusters_df, method_name){

  # Validate longitudinal consistency (same participant over time)
  validation_longitudinal <- clusters_df %>%
    mutate(nPID = sub("_W.*$", "", nSID)) %>%
    group_by(MAG_ID, species, nPID) %>%
    summarise(
      n_timepoints = n_distinct(nSID),
      n_clusters = n_distinct(strain_cluster),
      same_cluster = n_clusters == 1,
      .groups = "drop"
    ) %>%
    filter(n_timepoints > 1)

  # Build WGS strain mapping for each MAG/species
  wgs_mapping <- bind_rows(
    edges_weighted_new %>% select(MAG_ID, species, nSID = nSID_1, WGS_strain_num = WGS_strain_num_1),
    edges_weighted_new %>% select(MAG_ID, species, nSID = nSID_2, WGS_strain_num = WGS_strain_num_2)
  ) %>%
    distinct(MAG_ID, species, nSID, WGS_strain_num) %>%
    filter(!is.na(WGS_strain_num)) %>%
    mutate(WGS_strain_num = as.character(WGS_strain_num))

  # Join WGS labels to clusters
  clusters_with_WGS <- clusters_df %>%
    left_join(wgs_mapping, by = c("MAG_ID", "species", "nSID", "WGS_strain_num"))

  # Calculate WGS strain consistency (sensitivity: same WGS strain stays together)
  validation_WGS_sensitivity <- clusters_with_WGS %>%
    filter(!is.na(WGS_strain_num)) %>%
    group_by(MAG_ID, species, WGS_strain_num) %>%
    summarise(
      n_samples = n(),
      n_clusters = n_distinct(strain_cluster),
      same_cluster = n_clusters == 1,
      .groups = "drop"
    ) %>%
    filter(n_samples > 1)

  # Calculate WGS strain purity (specificity: each cluster contains only one WGS strain)
  validation_WGS_purity <- clusters_with_WGS %>%
    filter(!is.na(WGS_strain_num)) %>%
    group_by(MAG_ID, species, strain_cluster) %>%
    summarise(
      n_samples = n(),
      n_WGS_strains = n_distinct(WGS_strain_num),
      is_pure = n_WGS_strains == 1,
      .groups = "drop"
    ) %>%
    filter(n_samples > 1)

  # Aggregate validation metrics
  long_consistency <- if (nrow(validation_longitudinal) > 0) {
    mean(validation_longitudinal$same_cluster)
  } else {
    NA_real_
  }

  WGS_sensitivity <- if (nrow(validation_WGS_sensitivity) > 0) {
    mean(validation_WGS_sensitivity$same_cluster)
  } else {
    NA_real_
  }

  WGS_purity <- if (nrow(validation_WGS_purity) > 0) {
    mean(validation_WGS_purity$is_pure)
  } else {
    NA_real_
  }

  # Count total clusters
  n_total_clusters <- clusters_df %>%
    distinct(MAG_ID, species, strain_cluster) %>%
    nrow()

  # Return results
  tibble(
    method = method_name,
    n_total_clusters = n_total_clusters,
    longitudinal_consistency = long_consistency,
    WGS_strain_consistency = WGS_sensitivity,
    WGS_strain_purity = WGS_purity
  )

}


# Summarize clustering comparison metrics
comparison <- bind_rows(
  evaluate_clusters(strain_clusters_simple_val, "Connected Components"),
  evaluate_clusters(strain_clusters_louvain_val, "Louvain")
)


```

### 6.7 Supplementary Figure 7

```{r Supplementary Figure 7}

# Panel A: scaffold proportions
p_scaffolds <- ggplot(edges_weighted_old, aes(x = prop_scaffolds_high)) +
  geom_histogram(bins = 50, fill = "paleturquoise4", alpha = 0.7) +
  geom_vline(
    xintercept = c(0.8, 0.85, 0.9, 0.95, 1.0),
    color = "coral",
    linetype = "dashed"
  ) +
  labs(x = "Shared scaffold proportion (%)") +
  theme_minimal(base_size = 10)

# Panel B: MAG proportions
p_SGB <- ggplot(edges_weighted_old, aes(x = prop_MAG_scaffolds)) +
  geom_histogram(bins = 50, fill = "paleturquoise4", alpha = 0.7) +
  geom_vline(
    xintercept = c(0.5, 0.6, 0.7),
    color = "coral",
    linetype = "dashed"
  ) +
  labs(x = "Breadth (%)") +
  theme_minimal(base_size = 10)

# Select MAG to plot
mag_id <- "APPS0055PR_S_2_S113_concoct_bin45"

# Panel C: initial thresholds
pa <- plot_strain_network(
  instrain_old %>%
    filter(shared_strain == TRUE) %>%
    select(
      MAG_ID, species,
      ID_1, WGS_strain_num_1,
      ID_2, WGS_strain_num_2,
      Seq_1, Seq_2
    ),
  mag_id
) +
  labs(title = "Initial thresholds")

# Panel D: optimal thresholds and Louvain clustering
plot_strain_network_louvain <- function(dataframe, mag_id_to_plot, resolution = 0.3) {

  edges_mag <- dataframe %>%
    filter(MAG_ID == mag_id_to_plot) %>%
    select(
      ID_1, WGS_strain_num_1,
      ID_2, WGS_strain_num_2,
      Seq_1, Seq_2,
      median_popANI, prop_scaffolds_high
    )

  # Map mNGS IDs to WGS strain IDs
  mngs_to_wgs_map <- edges_mag %>%
    filter(
      ID_1 == ID_2,
      Seq_1 != Seq_2,
      (Seq_1 == "WGS" | Seq_2 == "WGS"),
      (Seq_1 == "mNGS" | Seq_2 == "mNGS")
    ) %>%
    {
      bind_rows(
        filter(., Seq_1 == "mNGS", Seq_2 == "WGS") %>%
          transmute(
            mngs_id = ID_1,
            wgs_id = paste(ID_2, WGS_strain_num_2, sep = "_")
          ),
        filter(., Seq_1 == "WGS", Seq_2 == "mNGS") %>%
          transmute(
            mngs_id = ID_2,
            wgs_id = paste(ID_1, WGS_strain_num_1, sep = "_")
          )
      )
    } %>%
    distinct()

  edges_mag <- edges_mag %>%
    left_join(
      mngs_to_wgs_map %>%
        rename(
          ID_1 = mngs_id,
          unified_ID_1_mapped = wgs_id
        ),
      by = "ID_1"
    ) %>%
    left_join(
      mngs_to_wgs_map %>%
        rename(
          ID_2 = mngs_id,
          unified_ID_2_mapped = wgs_id
        ),
      by = "ID_2"
    ) %>%
    mutate(
      unified_ID_1 = case_when(
        Seq_1 == "WGS" & !is.na(WGS_strain_num_1) ~
          paste(ID_1, WGS_strain_num_1, sep = "_"),
        Seq_1 == "mNGS" & !is.na(unified_ID_1_mapped) ~
          unified_ID_1_mapped,
        TRUE ~ ID_1
      ),
      unified_ID_2 = case_when(
        Seq_2 == "WGS" & !is.na(WGS_strain_num_2) ~
          paste(ID_2, WGS_strain_num_2, sep = "_"),
        Seq_2 == "mNGS" & !is.na(unified_ID_2_mapped) ~
          unified_ID_2_mapped,
        TRUE ~ ID_2
      )
    ) %>%
    filter(unified_ID_1 != unified_ID_2)

  if (nrow(edges_mag) == 0) return(NULL)

  g <- graph_from_data_frame(
    edges_mag[, c("unified_ID_1", "unified_ID_2")],
    directed = FALSE
  )

  E(g)$weight <- edges_mag$median_popANI *
    edges_mag$prop_scaffolds_high

  clust <- cluster_louvain(
    g,
    weights = E(g)$weight,
    resolution = resolution
  )

  V(g)$strain_cluster <- membership(clust)

  seq_info <- bind_rows(
    edges_mag %>%
      select(unified_ID = unified_ID_1, seq_type = Seq_1),
    edges_mag %>%
      select(unified_ID = unified_ID_2, seq_type = Seq_2)
  ) %>%
    group_by(unified_ID) %>%
    summarise(
      has_mNGS = any(seq_type == "mNGS"),
      has_WGS = any(seq_type == "WGS"),
      seq_label = case_when(
        has_mNGS & has_WGS ~ "Both",
        has_mNGS ~ "mNGS",
        has_WGS ~ "WGS"
      ),
      .groups = "drop"
    )

  V(g)$seq_type <- seq_info$seq_label[
    match(V(g)$name, seq_info$unified_ID)
  ]

  ggraph(as_tbl_graph(g), layout = "fr") +
    geom_edge_link(alpha = 0.5, color = "gray50") +
    geom_node_point(
      aes(color = factor(strain_cluster), shape = seq_type),
      size = 3
    ) +
    geom_node_text(
      aes(label = name),
      repel = TRUE,
      size = 2,
      max.overlaps = Inf
    ) +
    scale_color_manual(values = rev(hcl.colors(14, "Earth"))) +
    scale_shape_manual(
      values = c("mNGS" = 16, "WGS" = 17, "Both" = 15),
      name = "Sequencing Type"
    ) +
    guides(color = "none") +
    labs(title = "Optimal thresholds + Louvain") +
    theme_graph() +
    theme(
      legend.position = "right",
      plot.title.position = "plot"
    )
}

pc <- plot_strain_network_louvain(
  instrain_new %>%
    filter(shared_strain == TRUE),
  mag_id
)

# Assemble final supplementary figure
finalplot_supp <- (
  (p_scaffolds | p_SGB) /
  (
    (pa + theme(legend.position = "none")) |
    (pc + theme(legend.position = "none"))
  )
) +
  plot_layout(heights = c(0.4, 1)) +
  plot_annotation(
    tag_levels = list(c("a", "b", "c", "d"))
  ) &
  theme(
    plot.margin = margin(0, 0, 0, 4, "pt"),
    plot.tag = element_text(face = "bold", size = 12)
  )

finalplot_supp

```

### 6.8 Supplementary Figure 8

```{r main_instrain networks}


plot_strain_network <- function(dataframe, mag_id_to_plot) {

  # Filter to specific MAG
  edges_mag <- dataframe %>%
    filter(MAG_ID == mag_id_to_plot) %>%
    select(ID_1, WGS_strain_num_1, ID_2, WGS_strain_num_2, Seq_1, Seq_2)

  # Step 1: Find all mNGS-WGS pairs that should be merged
  # (same ID, one is mNGS, one is WGS, and they're connected by an edge)
  mngs_wgs_pairs <- edges_mag %>%
    filter(ID_1 == ID_2,  # Same patient/timepoint/source
           Seq_1 != Seq_2,  # One is mNGS, one is WGS
           (Seq_1 == "WGS" | Seq_2 == "WGS"),  # At least one is WGS
           (Seq_1 == "mNGS" | Seq_2 == "mNGS"))  # At least one is mNGS

  # Create mapping: mNGS ID -> WGS ID (with strain number)
  mngs_to_wgs_map <- bind_rows(
    mngs_wgs_pairs %>%
      filter(Seq_1 == "mNGS", Seq_2 == "WGS") %>%
      transmute(mngs_id = ID_1, wgs_id = paste(ID_2, WGS_strain_num_2, sep = "_")),
    mngs_wgs_pairs %>%
      filter(Seq_1 == "WGS", Seq_2 == "mNGS") %>%
      transmute(mngs_id = ID_2, wgs_id = paste(ID_1, WGS_strain_num_1, sep = "_"))
  ) %>%
    distinct()

  # Step 2: Create unified IDs
  ## 1.7 WGS analysis always gets strain appended
  ## 1.2 mNGS and MAG characterization gets mapped to WGS ID if it shares strain, otherwise keeps original
  edges_mag <- edges_mag %>%
      left_join(mngs_to_wgs_map %>% rename(ID_1 = mngs_id, unified_ID_1_mapped = wgs_id),
            by = "ID_1") %>%
  left_join(mngs_to_wgs_map %>% rename(ID_2 = mngs_id, unified_ID_2_mapped = wgs_id),
            by = "ID_2") %>%
  mutate(
    unified_ID_1 = case_when(
      Seq_1 == "WGS" & !is.na(WGS_strain_num_1) ~ paste(ID_1, WGS_strain_num_1, sep = "_"),
      Seq_1 == "mNGS" & !is.na(unified_ID_1_mapped) ~ unified_ID_1_mapped,
      TRUE ~ ID_1
    ),
    unified_ID_2 = case_when(
      Seq_2 == "WGS" & !is.na(WGS_strain_num_2) ~ paste(ID_2, WGS_strain_num_2, sep = "_"),
      Seq_2 == "mNGS" & !is.na(unified_ID_2_mapped) ~ unified_ID_2_mapped,
      TRUE ~ ID_2
    )
  )

  # Remove self-loops (edges where unified_ID_1 == unified_ID_2)
  edges_mag <- edges_mag %>%
    filter(unified_ID_1 != unified_ID_2)

  # Check if we have any edges left
  if (nrow(edges_mag) == 0) {
    return(NULL)  # No network to plot
  }

  # Create graph with unified IDs
  g <- graph_from_data_frame(edges_mag[, c("unified_ID_1", "unified_ID_2")],
                             directed = FALSE)

  # Find connected components (strain clusters)
  clusters <- components(g)

  # Add cluster membership as vertex attribute
  V(g)$strain_cluster <- clusters$membership

  # Add patient ID
  V(g)$patient <- sub("_T.*$", "", V(g)$name)

  # Add sequencing type information
  seq_info <- bind_rows(
    edges_mag %>% select(unified_ID = unified_ID_1, seq_type = Seq_1),
    edges_mag %>% select(unified_ID = unified_ID_2, seq_type = Seq_2)
  ) %>%
    group_by(unified_ID) %>%
    summarise(
      has_mNGS = any(seq_type == "mNGS"),
      has_WGS = any(seq_type == "WGS"),
      seq_label = case_when(
        has_mNGS & has_WGS ~ "Both",
        has_mNGS ~ "mNGS",
        has_WGS ~ "WGS"
      ),
      .groups = "drop"
    )

  V(g)$seq_type <- seq_info$seq_label[match(V(g)$name, seq_info$unified_ID)]

  # Convert to tidygraph for ggraph
  tg <- as_tbl_graph(g)

  # Get species name for this MAG
  species_name <- unique(dataframe$species[dataframe$MAG_ID == mag_id_to_plot])

  species_name <- ifelse(species_name == "Enterobacter hormaechei_C",
                       "Enterobacter hormaechei",
                       ifelse(species_name == "Enterococcus_B faecium",
                              "Enterococcus faecium",
                              species_name))

  # Plot
  p <- ggraph(tg, layout = 'fr') +
    geom_edge_link(aes(), alpha = 0.5, color = "gray50") +
    geom_node_point(aes(color = factor(strain_cluster),
                       shape = seq_type),
                   size = 3) +
    scale_color_manual(
      name = "Strain Cluster",
      values = rev(hcl.colors(11, "Earth"))  # Set to max you might need
      ) +
    scale_shape_manual(values = c("mNGS" = 16, "WGS" = 17, "Both" = 15),
                       name = "Sequencing Type") +
    labs(title = bquote(italic(.(species_name)))) +
    theme_graph() +
    theme(legend.position = "none",
          plot.title.position = "plot",
          plot.title = element_text(size=30))

  return(p)
}

instrain_new_df <- mNGS_instrain %>%
  select(MAG_ID, species, ID_1, WGS_strain_num_1, ID_2, WGS_strain_num_2, Seq_1, Seq_2)

# Get unique MAG IDs
ordered_MAG_ID <- instrain_new_df %>% select(MAG_ID, species) %>% distinct() %>%
  filter(species!="Pseudomonas aeruginosa") %>%
  arrange(species)

unique_mags_new <- ordered_MAG_ID$MAG_ID # ordered alphabetically, and excludes Pseudomonas since that's already in the paper

# Generate plots for all MAGs
plot_list <- list()

for (mag in unique_mags_new) {
  tryCatch({
    p <- plot_strain_network(instrain_new_df, mag)

    if (!is.null(p)) {
      plot_list[[mag]] <- p
    }

  }, error = function(e) {
    cat("Error with MAG", mag, ":", conditionMessage(e), "\n")
  })
}

# Arrange plots into a 3x3 panel
panel <- wrap_plots(plot_list, ncol = 3) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

# Temporarily restore one legend for collection
plot_list[[3]] <- plot_list[[3]] +
  theme(
    legend.position = "right",
    legend.text = element_text(size = 20),
    legend.title = element_text(size = 20)
  ) +
  guides(
    color = guide_legend(nrow = 2, direction = "horizontal", title.position = "top",
                         override.aes = list(size = 8)),
    shape = guide_legend(nrow = 1, direction = "horizontal", title.position = "top",
                         override.aes = list(size = 8))
  )

panel <- wrap_plots(plot_list, ncol = 3) +
  guide_area() +
  plot_layout(ncol = 3, guides = "collect") &
  theme(
    plot.background  = element_blank(),
    legend.text      = element_text(size = 20),
    legend.title     = element_text(size = 20, face = "bold"),
    legend.key.size  = unit(1.5, "cm")
  )

panel

```


## 7. WGS/MAG comparison

### 7.1 Supplementary Figure 9

```{r WGS vs MAG instrain comparison}
# Build the inStrain dataset

instrain1 <- mNGS_instrain %>%
  filter((Seq_1 == "WGS" | Seq_2 == "WGS") & shared_strain == TRUE) %>%
  select(species, sample_name_1, ID_1, Seq_1) %>%
  rename(
    sample_name = sample_name_1,
    ID = ID_1,
    Seq = Seq_1
  )

instrain2 <- mNGS_instrain %>%
  filter((Seq_1 == "WGS" | Seq_2 == "WGS") & shared_strain == TRUE) %>%
  select(species, sample_name_2, ID_2, Seq_2) %>%
  rename(
    sample_name = sample_name_2,
    ID = ID_2,
    Seq = Seq_2
  )

instrain_dx_final <- bind_rows(instrain1, instrain2) %>%
  filter(Seq == "mNGS") %>%
  mutate(
    nSID = str_extract(ID, "^[^_]+_[^_]+"),
    Source = ifelse(grepl("3279", sample_name), "stool", "PR"),
    value = 1
  ) %>%
  select(species, nSID, Source, value) %>%
  distinct() %>%
  pivot_wider(
    names_from = species,
    values_from = value,
    values_fill = 0
  ) %>%
  rename(
    SGBC.difficile = `Clostridioides difficile`,
    SGBK.pneumoniae = `Klebsiella pneumoniae`,
    SGBE.faecium = `Enterococcus_B faecium`,
    SGBP.aeruginosa = `Pseudomonas aeruginosa`,
    SGBE.faecalis = `Enterococcus faecalis`,
    SGBE.coli = `Escherichia coli`,
    SGBE.hormaechei = `Enterobacter hormaechei_C`
  ) %>%
  full_join(
    MAGs_SOI %>% select(mNGS_sample_name, nSID, Source),
    by = c("nSID", "Source")
  ) %>%
  mutate(across(starts_with("SGB"), ~replace_na(., 0)))


# Run inStrain analysis

run_instrain_analysis <- function(df, org_micro_name, org_col_name) {

  org_MAGs <- df %>%
    select(mNGS_sample_name, all_of(org_col_name)) %>%
    rename(MAG_detect = all_of(org_col_name)) %>%
    mutate(MAG_detect = as.integer(MAG_detect != 0))

  org_micro <- micro_cx %>%
    filter(MDRO == org_micro_name, Source == "PR") %>%
    distinct(nSID) %>%
    mutate(cx_detect = 1)

  org <- samples %>%
    filter(Collected == 1, Source == "PR") %>%
    full_join(
      mNGS_samples %>%
        select(mNGS_sample_name) %>%
        filter(grepl("3279", mNGS_sample_name)),
      by = c("MAG_sample" = "mNGS_sample_name")
    ) %>%
    left_join(org_MAGs, by = c("MAG_sample" = "mNGS_sample_name")) %>%
    left_join(org_micro, by = "nSID") %>%
    mutate(cx_detect = replace_na(cx_detect, 0)) %>%
    filter(!is.na(MAG_detect))

  twobytwo <- org %>%
    count(cx_detect, MAG_detect) %>%
    mutate(
      classification = case_when(
        cx_detect == 1 & MAG_detect == 1 ~ "Both +",
        cx_detect == 0 & MAG_detect == 0 ~ "Both -",
        cx_detect == 1 & MAG_detect == 0 ~ "WGS only +",
        cx_detect == 0 & MAG_detect == 1 ~ "SGB only +"
      )
    )

  get_n <- function(cx, mag) {
    x <- twobytwo %>%
      filter(cx_detect == cx, MAG_detect == mag) %>%
      pull(n)
    if (length(x) == 0) 0 else x
  }

  tp <- get_n(1, 1)
  fp <- get_n(0, 1)
  fn <- get_n(1, 0)
  tn <- get_n(0, 0)

  safe_binom <- function(x, n) {
    if (n == 0) return(list(estimate = NA, conf.int = c(NA, NA)))
    binom.test(x, n)
  }

  tests <- list(
    Sensitivity = safe_binom(tp, tp + fn),
    Specificity = safe_binom(tn, tn + fp),
    PPV         = safe_binom(tp, tp + fp),
    NPV         = safe_binom(tn, tn + fn)
  )

  perf <- tibble(
    Metric = names(tests),
    Estimate = sapply(tests, \(x) x$estimate),
    CI_lower = sapply(tests, \(x) x$conf.int[1]),
    CI_upper = sapply(tests, \(x) x$conf.int[2])
  )

  org_long <- org %>%
    pivot_longer(
      c(cx_detect, MAG_detect),
      names_to = "method",
      values_to = "detected"
    ) %>%
    mutate(
      method = factor(method, levels = c("cx_detect", "MAG_detect")),
      nPID_num = as.numeric(as.factor(nPID))
    )

  gee_p <- tryCatch({
    fit <- geeglm(
      detected ~ method,
      data = org_long,
      family = binomial,
      id = nPID_num,
      corstr = "independence"
    )
    summary(fit)$coefficients["methodMAG_detect", "Pr(>|W|)"]
  }, error = function(e) NA)

  list(
    perf = perf,
    twobytwo = twobytwo,
    gee_p = gee_p
  )
}


# Build inStrain plot

plot_instrain_combined <- function(results_list, title = "") {

  perf_all <- purrr::imap_dfr(
    results_list,
    ~.x$perf %>% mutate(Organism = .y)
  )

  concordance_all <- purrr::imap_dfr(
    results_list,
    ~.x$twobytwo %>% mutate(Organism = .y)
  )

  p_labels <- purrr::imap_dfr(
    results_list,
    ~tibble(Organism = .y, gee_p = .x$gee_p)
  )

  # Performance panel
  perf_combined <- perf_all %>%
    mutate(
      Organism = factor(Organism, levels = names(results_list)),
      Estimate_pct = Estimate * 100,
      CI_lower_pct = CI_lower * 100,
      CI_upper_pct = CI_upper * 100
    )

  metric_colors <- c(
    "Sensitivity" = "#4F6980FF",
    "Specificity" = "#F47942FF",
    "PPV"         = "#A2CEAAFF",
    "NPV"         = "#B66353FF"
  )

  pA <- ggplot(
    perf_combined,
    aes(x = Organism, y = Estimate_pct, fill = Metric)
  ) +
    geom_col(
      position = position_dodge(width = 0.75),
      width = 0.65
    ) +
    geom_errorbar(
      aes(ymin = CI_lower_pct, ymax = CI_upper_pct),
      position = position_dodge(width = 0.75),
      width = 0.25
    ) +
    scale_y_continuous(
      limits = c(0, 100),
      labels = scales::label_percent(scale = 1),
      breaks = seq(0, 100, 25)
    ) +
    scale_fill_manual(values = metric_colors) +
    labs(x = NULL, y = "Estimate (%)", fill = "Performance metrics") +
    theme_classic(base_size = 13) +
    ggtitle(title)

  # Concordance panel
  concordance_combined <- concordance_all %>%
    left_join(
      concordance_all %>%
        group_by(Organism) %>%
        summarise(total = sum(n), .groups = "drop"),
      by = "Organism"
    ) %>%
    left_join(p_labels, by = "Organism") %>%
    mutate(
      Organism = factor(Organism, levels = names(results_list)),
      prop = n / total * 100
    )

  annot_df <- concordance_combined %>%
    distinct(Organism, gee_p) %>%
    mutate(
      label = ifelse(
        is.na(gee_p), "N/A",
        ifelse(gee_p < 0.01, "p <0.01", sprintf("p = %.2f", gee_p))
      )
    )

  pB <- ggplot(
    concordance_combined,
    aes(x = Organism, y = prop, fill = classification)
  ) +
    geom_col(width = 0.5) +
    geom_text(
      data = annot_df,
      aes(x = Organism, y = 102, label = label),
      inherit.aes = FALSE,
      size = 4
    ) +
    scale_fill_manual(
      values = c(
        "WGS only +" = "#F47942FF",
        "Both +"     = "#638B66FF",
        "SGB only +" = "#4F6980FF",
        "Both -"     = "grey60"
      ),
      labels = c(
        "WGS only +" = "culture only +",
        "Both +"     = "Both +",
        "SGB only +" = "mNGS only +",
        "Both -"     = "Both −"
      )
    ) +
    scale_y_continuous(
      limits = c(0, 115),
      labels = scales::label_percent(scale = 1),
      breaks = seq(0, 100, 25)
    ) +
    labs(x = NULL, y = "Proportion of samples (%)", fill = "Detection method") +
    theme_classic(base_size = 13) +
    ggtitle(title)

  list(pA = pA, pB = pB)
}


# Assemble final figure

all_results_edges <- list(
  "C. difficile"  = run_instrain_analysis(instrain_dx_final, "C. difficile",  "SGBC.difficile"),
  "E. faecium"    = run_instrain_analysis(instrain_dx_final, "E. faecium",    "SGBE.faecium"),
  "E. faecalis"   = run_instrain_analysis(instrain_dx_final, "E. faecalis",   "SGBE.faecalis"),
  "K. pneumoniae" = run_instrain_analysis(instrain_dx_final, "K. pneumoniae", "SGBK.pneumoniae"),
  "P. aeruginosa" = run_instrain_analysis(instrain_dx_final, "P. aeruginosa", "SGBP.aeruginosa"),
  "E. coli"       = run_instrain_analysis(instrain_dx_final, "E. coli",       "SGBE.coli"),
  "E. hormaechei" = run_instrain_analysis(instrain_dx_final, "E. hormaechei", "SGBE.hormaechei")
)

instrain_fig_edges <- plot_instrain_combined(
  all_results_edges,
  title = ""
)

instrain_plot_top <-
  instrain_fig_edges$pA /
  instrain_fig_edges$pB +
  plot_layout(heights = c(1, 1)) +
  plot_annotation(tag_levels = "a") &
  theme(plot.tag = element_text(face = "bold", size = 24))

instrain_plot_top

```


## 8. Risk factor analysis

### 8.1 Data preparation

```{r data preparation}

# Build same-timepoint P×P dyads
pp_temp <- SSS %>%
  filter(
    str_starts(nPID_1, "P"),
    str_starts(nPID_2, "P"),
    !same_patient,
    nTID_1 == nTID_2
  ) %>%
  rename(week = nTID_1) %>%
  select(-nTID_2) %>%
  mutate(
    pid_a = if_else(nPID_1 < nPID_2, nPID_1, nPID_2),
    pid_b = if_else(nPID_1 < nPID_2, nPID_2, nPID_1)
  ) %>%
  select(pid_a, pid_b, week, n_MAGs_compared, sharing_rate) %>%
  left_join(week_dates, by = "week")

# Count intervals contributed by each patient
dyad_intervals_temp <- pp_temp %>%
  group_by(pid_a, pid_b) %>%
  arrange(week_num, .by_group = TRUE) %>%
  mutate(week_t2 = lead(week)) %>%
  filter(!is.na(week_t2)) %>%
  ungroup()

patient_interval_counts <- bind_rows(
  dyad_intervals_temp %>% count(pid = pid_a),
  dyad_intervals_temp %>% count(pid = pid_b)
) %>%
  group_by(pid) %>%
  summarise(
    n_intervals = sum(n),
    .groups = "drop"
  )

# Order dyads so pid_a has more intervals
pp_same <- SSS %>%
  filter(
    str_starts(nPID_1, "P"),
    str_starts(nPID_2, "P"),
    !same_patient,
    nTID_1 == nTID_2
  ) %>%
  rename(week = nTID_1) %>%
  select(-nTID_2) %>%
  left_join(
    patient_interval_counts %>%
      rename(nPID_1 = pid, n1 = n_intervals),
    by = "nPID_1"
  ) %>%
  left_join(
    patient_interval_counts %>%
      rename(nPID_2 = pid, n2 = n_intervals),
    by = "nPID_2"
  ) %>%
  mutate(
    pid_a = case_when(
      n1 > n2         ~ nPID_1,
      n2 > n1         ~ nPID_2,
      nPID_1 < nPID_2 ~ nPID_1,
      TRUE            ~ nPID_2
    ),
    pid_b = case_when(
      n1 > n2         ~ nPID_2,
      n2 > n1         ~ nPID_1,
      nPID_1 < nPID_2 ~ nPID_2,
      TRUE            ~ nPID_1
    )
  ) %>%
  select(
    pid_a, pid_b, week,
    n_MAGs_compared,
    sharing_rate
  ) %>%
  left_join(week_dates, by = "week") %>%
  mutate(
    log_n_MAGs = log(n_MAGs_compared)
  )


# Create consecutive T1 → T2 dyad intervals
dyad_intervals <- pp_same %>%
  group_by(pid_a, pid_b) %>%
  arrange(week_num, .by_group = TRUE) %>%
  mutate(
    week_t2         = lead(week),
    sharing_rate_t2 = lead(sharing_rate),
    log_n_MAGs_t2   = lead(log_n_MAGs),
    interval_start  = w_start,
    interval_end    = lead(w_end)
  ) %>%
  filter(!is.na(week_t2)) %>%
  ungroup() %>%
  rename(
    week_t1         = week,
    sharing_rate_t1 = sharing_rate
  )


# Create shared model dataset
model_data <- dyad_intervals %>%
  select(
    pid_a,
    pid_b,
    week_t1,
    interval_start,
    interval_end,
    sharing_rate_t1,
    sharing_rate_t2,
    log_n_MAGs_t2
  ) %>%
  mutate(
    sharing_rate_adj = case_when(
      sharing_rate_t2 == 0 ~ 0.0001,
      sharing_rate_t2 == 1 ~ 0.9999,
      TRUE                 ~ sharing_rate_t2
    )
  )


# Create daily room-occupancy data
rooms_daily <- rooms_raw %>%
  rowwise() %>%
  mutate(
    date = list(seq(AdmDate, DcDate, by = "day"))
  ) %>%
  unnest(date) %>%
  ungroup() %>%
  select(nPID, date, Room, Floor)
```


### 8.2 Figure 5

```{r provider crossings analysis}

# Link provider visits to patients
provider_visits <- providers_raw %>%
  left_join(
    rooms_daily %>% rename(room = Room),
    by = c("room", "event_date" = "date")
  ) %>%
  filter(
    !is.na(nPID),
    str_starts(nPID, "P")
  ) %>%
  select(
    clinician_id,
    group_classification,
    event_date,
    timestamp,
    nPID
  ) %>%
  mutate(
    datetime = ymd_hms(paste(event_date, timestamp))
  ) %>%
  distinct(
    nPID,
    clinician_id,
    group_classification,
    datetime
  )


# Create patient-specific visit records
all_visits_a <- dyad_intervals %>%
  select(
    pid_a, pid_b, week_t1,
    interval_start, interval_end
  ) %>%
  left_join(
    provider_visits %>% rename(pid_a = nPID),
    by = "pid_a",
    relationship = "many-to-many"
  ) %>%
  filter(
    datetime >= ymd(interval_start),
    datetime <= ymd(interval_end) + hours(23) + minutes(59)
  ) %>%
  mutate(patient = "a")

all_visits_b <- dyad_intervals %>%
  select(
    pid_a, pid_b, week_t1,
    interval_start, interval_end
  ) %>%
  left_join(
    provider_visits %>% rename(pid_b = nPID),
    by = "pid_b",
    relationship = "many-to-many"
  ) %>%
  filter(
    datetime >= ymd(interval_start),
    datetime <= ymd(interval_end) + hours(23) + minutes(59)
  ) %>%
  mutate(patient = "b")


# Identify clinician crossings between patients
crossing_features <- bind_rows(
  all_visits_a,
  all_visits_b
) %>%
  group_by(
    pid_a,
    pid_b,
    week_t1,
    clinician_id
  ) %>%
  filter(n_distinct(patient) == 2) %>%
  arrange(datetime, .by_group = TRUE) %>%
  mutate(
    crossed = patient != lag(patient, default = first(patient))
  ) %>%
  summarise(
    n_crossings = sum(crossed),
    group_classification = first(group_classification),
    .groups = "drop"
  )


# Aggregate crossings and normalize by interval length
crossing_summary <- crossing_features %>%
  group_by(pid_a, pid_b, week_t1) %>%
  summarise(
    total_crossings =
      sum(n_crossings),

    n_nurses_crossings =
      sum(
        n_crossings[
          group_classification == "Nurses (RN)"
        ],
        na.rm = TRUE
      ),

    n_rt_crossings =
      sum(
        n_crossings[
          group_classification ==
            "Respiratory Therapists (RT)"
        ],
        na.rm = TRUE
      ),

    n_techs_crossings =
      sum(
        n_crossings[
          group_classification == "Techs"
        ],
        na.rm = TRUE
      ),

    n_evs_crossings =
      sum(
        n_crossings[
          group_classification %in% c(
            "Environmental Services (EVS)",
            "Facility Services",
            "Food Services"
          )
        ],
        na.rm = TRUE
      ),

    n_therapist_crossings =
      sum(
        n_crossings[
          group_classification == "Therapists"
        ],
        na.rm = TRUE
      ),

    n_physician_crossings =
      sum(
        n_crossings[
          group_classification == "Physicians"
        ],
        na.rm = TRUE
      ),

    .groups = "drop"
  ) %>%
  left_join(
    dyad_intervals %>%
      select(
        pid_a,
        pid_b,
        week_t1,
        interval_start,
        interval_end
      ) %>%
      mutate(
        interval_days =
          as.numeric(interval_end - interval_start)
      ),
    by = c("pid_a", "pid_b", "week_t1")
  ) %>%
  mutate(
    crossings_per_day =
      total_crossings / interval_days,

    nurse_crossings_per_day =
      n_nurses_crossings / interval_days,

    rt_crossings_per_day =
      n_rt_crossings / interval_days,

    techs_crossings_per_day =
      n_techs_crossings / interval_days,

    evs_crossings_per_day =
      n_evs_crossings / interval_days,

    therapist_crossings_per_day =
      n_therapist_crossings / interval_days,

    physician_crossings_per_day =
      n_physician_crossings / interval_days
  )


# Add crossing predictors to the model dataset
model_data_crossings <- model_data %>%
  left_join(
    crossing_summary %>%
      select(
        pid_a,
        pid_b,
        week_t1,
        crossings_per_day,
        nurse_crossings_per_day,
        rt_crossings_per_day,
        techs_crossings_per_day,
        evs_crossings_per_day,
        therapist_crossings_per_day,
        physician_crossings_per_day
      ),
    by = c("pid_a", "pid_b", "week_t1")
  ) %>%
  mutate(
    across(
      ends_with("_per_day"),
      ~ replace_na(., 0)
    )
  )


# Fit crossing models
run_model <- function(formula, data) {
  glmmTMB(
    formula,
    ziformula = ~1,
    family = beta_family(link = "logit"),
    data = data
  )
}

extract_results <- function(fit, label) {

  coefs <- fixef(fit)$cond
  cis   <- confint(fit, parm = "beta_", method = "Wald")
  wald  <- car::Anova(fit)

  coef_name <- names(coefs)[3]
  ci_name   <- paste0("cond.", coef_name)

  tibble(
    label   = label,
    OR      = exp(coefs[coef_name]),
    CI_low  = exp(cis[ci_name, "2.5 %"]),
    CI_high = exp(cis[ci_name, "97.5 %"]),
    p       = wald[2, "Pr(>Chisq)"]
  )
}


fitc1 <- run_model(
  sharing_rate_adj ~
    sharing_rate_t1 +
    crossings_per_day +
    offset(log_n_MAGs_t2) +
    (1 | pid_a),
  model_data_crossings
)

fitc2 <- run_model(
  sharing_rate_adj ~
    sharing_rate_t1 +
    nurse_crossings_per_day +
    offset(log_n_MAGs_t2) +
    (1 | pid_a) + (1 | pid_b),
  model_data_crossings
)

fitc3 <- run_model(
  sharing_rate_adj ~
    sharing_rate_t1 +
    rt_crossings_per_day +
    offset(log_n_MAGs_t2) +
    (1 | pid_a) + (1 | pid_b),
  model_data_crossings
)

fitc4 <- run_model(
  sharing_rate_adj ~
    sharing_rate_t1 +
    techs_crossings_per_day +
    offset(log_n_MAGs_t2) +
    (1 | pid_a) + (1 | pid_b),
  model_data_crossings
)

fitc5 <- run_model(
  sharing_rate_adj ~
    sharing_rate_t1 +
    evs_crossings_per_day +
    offset(log_n_MAGs_t2) +
    (1 | pid_a) + (1 | pid_b),
  model_data_crossings
)

fitc6 <- run_model(
  sharing_rate_adj ~
    sharing_rate_t1 +
    therapist_crossings_per_day +
    offset(log_n_MAGs_t2) +
    (1 | pid_a) + (1 | pid_b),
  model_data_crossings
)

fitc7 <- run_model(
  sharing_rate_adj ~
    sharing_rate_t1 +
    physician_crossings_per_day +
    offset(log_n_MAGs_t2) +
    (1 | pid_a) + (1 | pid_b),
  model_data_crossings
)


forest_crossings <- bind_rows(
  extract_results(fitc1, "All staff"),
  extract_results(fitc2, "Nurses"),
  extract_results(fitc3, "Respiratory therapists"),
  extract_results(fitc4, "Techs"),
  extract_results(fitc5, "Support services"),
  extract_results(fitc6, "Therapists"),
  extract_results(fitc7, "Physicians")
) %>%
  mutate(
    sig = p < 0.05,
    label = factor(label, levels = rev(label))
  )


# Create forest plot
forest_plot <- ggplot(
  forest_crossings,
  aes(y = label)
) +
  geom_vline(
    xintercept = 1,
    linetype = "dashed",
    color = "grey60",
    linewidth = 0.4
  ) +
  geom_errorbarh(
    aes(xmin = CI_low, xmax = CI_high, color = sig),
    height = 0.2,
    linewidth = 0.6,
    show.legend = FALSE
  ) +
  geom_point(
    aes(x = OR, color = sig, shape = sig),
    size = 2.8
  ) +
  geom_text(
    aes(
      x = max(CI_high, na.rm = TRUE) * 1.05,
      label = paste0(
        sprintf(
          "OR: %.2f (95%% CI: %.2f–%.2f)",
          OR, CI_low, CI_high
        ),
        ifelse(
          p < 0.001,
          ", p<0.01",
          paste0(", p=", sprintf("%.2f", p))
        )
      ),
      color = sig
    ),
    hjust = 0,
    size = 3,
    show.legend = FALSE
  ) +
  scale_x_log10(
    breaks = c(0.5, 1, 2),
    labels = c("0.5", "1.0", "2.0"),
    limits = c(0.1, 40)
  ) +
  scale_color_manual(
    values = c(
      "TRUE" = "#185FA5",
      "FALSE" = "#888780"
    ),
    labels = c(
      "TRUE" = "p < 0.05",
      "FALSE" = "p ≥ 0.05"
    ),
    name = NULL
  ) +
  scale_shape_manual(
    values = c(
      "TRUE" = 18,
      "FALSE" = 16
    ),
    labels = c(
      "TRUE" = "p < 0.05",
      "FALSE" = "p ≥ 0.05"
    ),
    name = NULL
  ) +
  labs(
    x = "\nOdds ratio (log scale)",
    y = "Number of interpatient crossings\n"
  ) +
  theme_mine() +
  theme(
    legend.position = "bottom",
    axis.text.y = element_text(size = 10),
    panel.grid.major.x = element_line(
      color = "grey90",
      linewidth = 0.3
    ),
    panel.grid.major.y = element_blank()
  )

forest_plot
```


### 8.3 Additional risk factor analysis

```{r HAND HYGIENE + SAME-FLOOR ANALYSIS}

# Link provider events to patients with hand-hygiene data
provider_patient <- providers_raw %>%
  left_join(
    rooms_daily %>% rename(room = Room),
    by = c("room", "event_date" = "date")
  ) %>%
  filter(
    !is.na(nPID),
    str_starts(nPID, "P")
  ) %>%
  select(
    clinician_id,
    group_classification,
    event_date,
    is_performant,
    nPID
  )


# Calculate hand-hygiene non-performance among bridging staff
visits_a_hh <- dyad_intervals %>%
  select(
    pid_a, pid_b, week_t1,
    interval_start, interval_end
  ) %>%
  left_join(
    provider_patient %>% rename(pid_a = nPID),
    by = "pid_a",
    relationship = "many-to-many"
  ) %>%
  filter(
    event_date >= interval_start,
    event_date <= interval_end
  )

visits_b_hh <- dyad_intervals %>%
  select(
    pid_a, pid_b, week_t1,
    interval_start, interval_end
  ) %>%
  left_join(
    provider_patient %>% rename(pid_b = nPID),
    by = "pid_b",
    relationship = "many-to-many"
  ) %>%
  filter(
    event_date >= interval_start,
    event_date <= interval_end
  ) %>%
  select(
    pid_a, pid_b, week_t1, clinician_id
  )

bridging_hh <- visits_a_hh %>%
  inner_join(
    visits_b_hh,
    by = c(
      "pid_a",
      "pid_b",
      "week_t1",
      "clinician_id"
    ),
    relationship = "many-to-many"
  ) %>%
  group_by(pid_a, pid_b, week_t1) %>%
  summarise(
    bridge_nonperf_rate =
      mean(is_performant == "N", na.rm = TRUE),
    .groups = "drop"
  )


# Calculate same-floor exposure
floor_interval <- dyad_intervals %>%
  select(
    pid_a, pid_b, week_t1,
    interval_start, interval_end
  ) %>%
  left_join(
    rooms_daily %>%
      select(nPID, date, Floor) %>%
      rename(
        pid_a = nPID,
        floor_a = Floor
      ),
    by = "pid_a",
    relationship = "many-to-many"
  ) %>%
  filter(
    date >= interval_start,
    date <= interval_end
  ) %>%
  left_join(
    rooms_daily %>%
      select(nPID, date, Floor) %>%
      rename(
        pid_b = nPID,
        floor_b = Floor
      ),
    by = c("pid_b", "date")
  ) %>%
  group_by(pid_a, pid_b, week_t1) %>%
  summarise(
    days_both_present = sum(!is.na(floor_b)),
    days_same_floor   = sum(floor_a == floor_b, na.rm = TRUE),
    same_floor =
      days_both_present > 0 &
      days_same_floor / days_both_present > 0.5,
    .groups = "drop"
  )


# Add hand-hygiene and floor predictors to the model dataset
model_data_hh_floor <- model_data %>%
  left_join(
    bridging_hh,
    by = c("pid_a", "pid_b", "week_t1")
  ) %>%
  left_join(
    floor_interval %>%
      select(pid_a, pid_b, week_t1, same_floor),
    by = c("pid_a", "pid_b", "week_t1")
  ) %>%
  mutate(
    same_floor = factor(
      same_floor,
      levels = c(FALSE, TRUE),
      labels = c("Different floor", "Same floor")
    )
  )


# Fit crossing models
fit1 <- run_model(
  sharing_rate_adj ~
    sharing_rate_t1 +
    bridge_nonperf_rate +
    offset(log_n_MAGs_t2) +
    (1 | pid_a) + (1 | pid_b),
  model_data_hh_floor %>%
    filter(!is.na(bridge_nonperf_rate))
)

fit2 <- run_model(
  sharing_rate_adj ~
    sharing_rate_t1 +
    same_floor +
    offset(log_n_MAGs_t2) +
    (1 | pid_a) + (1 | pid_b),
  model_data_hh_floor %>%
    filter(!is.na(same_floor))
)


# Extract hand-hygiene and same-floor results
hh_result <- extract_results(
  fit1,
  "Hand hygiene non-compliance"
)

floor_result <- extract_results(
  fit2,
  "Same floor"
)

hh_result
floor_result
```
